---
title: "Sauerkraut study - Descriptive Bloodmarker Analysis"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

```{r packages}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(SummarizedExperiment) # microbiome analysis
library(mia)                  # microbiome analysis

library(dplyr)     # data handling
library(tidyr)     # data transformation
library(ggplot2)   # data visualization
library(patchwork) # grid of ggplot2

# set ggplot2 theme
theme_set(
  theme_minimal() +
    theme(plot.title       = element_text(hjust = 0.5),
          plot.subtitle    = element_text(hjust = 0.5),
          panel.grid.minor = element_blank(),
          plot.background  = element_rect(fill = "white", color = "white"))
)
```

# Data preparation

```{r set file paths}
define_dataPaths()
```

```{r read data}
#| warning: false

tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")

dat <- tse %>% 
  colData() %>% 
  as.data.frame()
```

```{r lookup vectors}
metabolic_markers    <- c("Fructosamine" = "fructosamine",
                          "Glucose"      = "glucose",
                          "Insulin"      = "insulin")
inflammation_markers <- c("IL-6"    = "il6",
                          "hsCRP"   = "scrp",
                          "sTNFR-2" = "tnfr2",
                          "sRAGE"   = "rage")
leakyGut_markers     <- c("LBP"     = "lbp",
                          "FABP-2"  = "fabp2",
                          "Zonulin" = "zonulin")
SCFA_bloodMarkers    <- c("Acetic acid (Blood)"          = "blood_aceticAcid",
                          "Propionic acid (Blood)"       = "blood_propionicAcid",
                          "Butyric acid (Blood)"         = "blood_butyricAcid",
                          "Valeric acid (Blood)"         = "blood_valericAcid",
                          "Hexanoic acid (Blood)"        = "blood_hexanoicAcid")
SCFA_stoolMarkers    <- c("Acetic acid (Stool)"          = "stool_aceticAcid",
                          "Propionic acid (Stool)"       = "stool_propionicAcid",
                          "Butyric acid (Stool)"         = "stool_butyricAcid",
                          "Valeric acid (Stool)"         = "stool_valericAcid",
                          "Hexanoic acid (Stool)"        = "stool_hexanoicAcid")
BCFA_bloodMarkers    <- c("Iso-butyric acid (Blood)"     = "blood_isobutyricAcid",
                          "Iso-valeric acid (Blood)"     = "blood_isovalericAcid",
                          "2-Methylbutyric acid (Blood)" = "blood_methylbutyricAcid")
BCFA_stoolMarkers    <- c("Iso-butyric acid (Stool)"     = "stool_isobutyricAcid",
                          "Iso-valeric acid (Stool)"     = "stool_isovalericAcid",
                          "2-Methylbutyric acid (Stool)" = "stool_methylbutyricAcid")

marker_units         <- c("Glucose"                      = "mg/dl",
                          "Fructosamine"                 = "µmol/l",
                          "Insulin"                      = "pmol/l",
                          "sTNFR-2"                      = "pg/ml",
                          "IL-6"                         = "pg/ml",
                          "hsCRP"                        = "mg/l",
                          "sRAGE"                        = "pg/ml",
                          "LBP"                          = "ng/ml",
                          "FABP-2"                       = "pg/ml",
                          "Zonulin"                      = "ng/ml",
                          "2-Methylbutyric acid (Blood)" = "ng/ml",
                          "Acetic acid (Blood)"          = "ng/ml",
                          "Butyric acid (Blood)"         = "ng/ml",
                          "Hexanoic acid (Blood)"        = "ng/ml",
                          "Iso-butyric acid (Blood)"     = "ng/ml",
                          "Iso-valeric acid (Blood)"     = "ng/ml",
                          "Propionic acid (Blood)"       = "ng/ml",
                          "Valeric acid (Blood)"         = "ng/ml",
                          "2-Methylbutyric acid (Stool)" = "µg/g",
                          "Acetic acid (Stool)"          = "µg/g",
                          "Butyric acid (Stool)"         = "µg/g",
                          "Hexanoic acid (Stool)"        = "µg/g",
                          "Iso-butyric acid (Stool)"     = "µg/g",
                          "Iso-valeric acid (Stool)"     = "µg/g",
                          "Propionic acid (Stool)"       = "µg/g",
                          "Valeric acid (Stool)"         = "µg/g")

col_vector  <- c("metabolites"  = "#E4F5E0",
                 "inflammation" = "#FCEEF3",
                 "leaky gut"    = "#FDEED9",
                 "scfa (blood)" = "#DBEAFB",
                 "scfa (stool)" = "#DBEAFB",
                 "bcfa (blood)" = "#EBE1F4",
                 "bcfa (stool)" = "#EBE1F4")
col_vector2 <- c("metabolites"  = "#C8EBC1",
                 "inflammation" = "#F5CCDC",
                 "leaky gut"    = "#FADEB3",
                 "scfa (blood)" = "#B7D4F6",
                 "scfa (stool)" = "#B7D4F6",
                 "bcfa (blood)" = "#D6C3E9",
                 "bcfa (stool)" = "#D6C3E9")

col_vector_withoutSCFAs  <- c("metabolites"  = "#E4F5E0",
                              "inflammation" = "#FCEEF3",
                              "leaky gut"    = "#DBEAFB")
col_vector_withoutSCFAs2 <- c("metabolites"  = "#C8EBC1",
                              "inflammation" = "#F5CCDC",
                              "leaky gut"    = "#B7D4F6")
```

```{r create marker dataset}
# restrict dataset to relevant variables and change marker data to long format
dat_markers <- dat %>% 
  select(participant_id, timepoint, group, treatment, intervention, inter_treat,
         gender, age, bmi_t0,
         glucose, fructosamine, insulin,
         tnfr2, il6, scrp, rage,
         lbp, fabp2, zonulin,
         contains("Acid")) %>% 
  pivot_longer(cols = all_of(c("glucose", "fructosamine", "insulin", "tnfr2", "il6", "scrp", "rage", "lbp", "fabp2", "zonulin",
                               "blood_methylbutyricAcid", "blood_aceticAcid", "blood_butyricAcid", "blood_hexanoicAcid",
                               "blood_isobutyricAcid", "blood_isovalericAcid", "blood_propionicAcid", "blood_valericAcid",
                               "stool_methylbutyricAcid", "stool_aceticAcid", "stool_butyricAcid", "stool_hexanoicAcid",
                               "stool_isobutyricAcid", "stool_isovalericAcid", "stool_propionicAcid", "stool_valericAcid")),
               names_to = "marker")

# add marker group variable
dat_markers <- dat_markers %>% 
  mutate(marker_group = case_when(marker %in% metabolic_markers    ~ "Metabolites",
                                  marker %in% inflammation_markers ~ "Inflammation",
                                  marker %in% leakyGut_markers     ~ "Leaky gut",
                                  marker %in% SCFA_bloodMarkers    ~ "SCFA (blood)",
                                  marker %in% SCFA_stoolMarkers    ~ "SCFA (stool)",
                                  marker %in% BCFA_bloodMarkers    ~ "BCFA (blood)",
                                  TRUE                             ~ "BCFA (stool)"),
         marker = case_when(marker_group == "Metabolites"  ~ names(metabolic_markers)[   match(marker, metabolic_markers)],
                            marker_group == "Inflammation" ~ names(inflammation_markers)[match(marker, inflammation_markers)],
                            marker_group == "Leaky gut"    ~ names(leakyGut_markers)[    match(marker, leakyGut_markers)],
                            marker_group == "SCFA (blood)" ~ names(SCFA_bloodMarkers)[    match(marker, SCFA_bloodMarkers)],
                            marker_group == "SCFA (stool)" ~ names(SCFA_stoolMarkers)[    match(marker, SCFA_stoolMarkers)],
                            marker_group == "BCFA (blood)" ~ names(BCFA_bloodMarkers)[    match(marker, BCFA_bloodMarkers)],
                            TRUE                           ~ names(BCFA_stoolMarkers)[    match(marker, BCFA_stoolMarkers)]))
```

```{r marker development}
# compute if a person's marker values fell or sunk
dat_markersDiff <- dat_markers %>% 
  filter(intervention != "FollowUp") %>% 
  arrange(participant_id, marker, intervention, treatment) %>% 
  group_by(participant_id, marker, intervention) %>% 
  filter(n() == 2) %>% 
  mutate(value_diff = diff(value)) %>% 
  mutate(value_diffCat = ifelse(value_diff > 0, "positive", "negative"))
# add this information to the main dataset
ids_dat <- paste(dat_markers$participant_id,     dat_markers$marker,     dat_markers$intervention)
ids_new <- paste(dat_markersDiff$participant_id, dat_markersDiff$marker, dat_markersDiff$intervention)
dat_markers <- dat_markers %>%
  mutate(value_diff    = dat_markersDiff$value_diff[   match(ids_dat, ids_new)],
         value_diffCat = dat_markersDiff$value_diffCat[match(ids_dat, ids_new)])
```


# Plot blood markers

```{r}
# helper function
plot_oneMarker <- function(markerGroup_name, marker_name, y_log10 = FALSE,
                           col_vec = col_vector, col_vec2 = col_vector2) {
  gg <- dat_markers %>% 
    filter(marker_group == markerGroup_name,
           marker       == marker_name) %>% 
    ggplot(aes(x = treatment, y = value)) +
    geom_boxplot(fill = col_vec2[tolower(markerGroup_name)], width = 0.5) +
    geom_line(aes(group = participant_id, col = value_diffCat), alpha = .3) +
    facet_grid(cols = vars(intervention), rows = vars(marker), scales = "free", space = "free_x") +
    theme(axis.title.x       = element_blank(),
          plot.title         = element_text(hjust = 0.5),
          strip.background.y = element_rect(fill = col_vec[tolower(markerGroup_name)], color = col_vec[tolower(markerGroup_name)]),
          strip.background.x = element_rect(fill = "gray95", color = "gray95"),
          legend.position    = "none")
  
  gg <- if (y_log10) {
    gg + scale_y_log10(paste0(marker_units[marker_name], ", on log10 scale"))
  } else {
    gg + ylab(marker_units[marker_name])
  }
  
  gg
}
```


## Metabolic markers

```{r plot metabolic markers}
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 6

gg_frc <- plot_oneMarker("Metabolites", "Fructosamine") + theme(axis.text.x  = element_blank())
gg_glc <- plot_oneMarker("Metabolites", "Glucose")      + theme(axis.text.x  = element_blank())
gg_ins <- plot_oneMarker("Metabolites", "Insulin")      + theme(strip.text.x = element_blank())

(gg_frc + gg_glc) / (gg_ins + ggplot())
```


## Inflammation markers

Note: The following y axes are logarithmized, but the labels along the axes are
original, unlogarithmized values.

```{r plot inflammation markers}
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 6

gg_il6   <- plot_oneMarker("Inflammation", "IL-6",   y_log10 = TRUE) + theme(axis.text.x  = element_blank())
gg_hscrp <- plot_oneMarker("Inflammation", "hsCRP", y_log10 = TRUE) + theme(axis.text.x  = element_blank())
gg_tnfr2 <- plot_oneMarker("Inflammation", "sTNFR-2", y_log10 = TRUE) + theme(strip.text.x = element_blank())
gg_rage  <- plot_oneMarker("Inflammation", "sRAGE",  y_log10 = TRUE) + theme(strip.text.x = element_blank())

(gg_il6 + gg_hscrp) / (gg_tnfr2 + gg_rage)
```

## Leaky gut markers

```{r plot leaky gut markers}
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 6

gg_lbp     <- plot_oneMarker("Leaky gut", "LBP")     + theme(axis.text.x  = element_blank())
gg_fabp2   <- plot_oneMarker("Leaky gut", "FABP-2")   + theme(axis.text.x  = element_blank())
gg_zonulin <- plot_oneMarker("Leaky gut", "Zonulin") + theme(strip.text.x = element_blank())

(gg_lbp + gg_fabp2) / (gg_zonulin + ggplot())
```


## Short / Branch chain fatty acids (Blood measurements)

```{r plot blood SCFAs}
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 6

gg_aca     <- plot_oneMarker("SCFA (blood)", "Acetic acid (Blood)")          + theme(axis.text.x  = element_blank())
gg_ppa     <- plot_oneMarker("SCFA (blood)", "Propionic acid (Blood)")       + theme(axis.text.x  = element_blank())
gg_bta     <- plot_oneMarker("SCFA (blood)", "Butyric acid (Blood)")         + theme(strip.text.x = element_blank())
gg_vla     <- plot_oneMarker("SCFA (blood)", "Valeric acid (Blood)")         + theme(strip.text.x = element_blank())
gg_hxa     <- plot_oneMarker("SCFA (blood)", "Hexanoic acid (Blood)")        + theme(axis.text.x  = element_blank())

gg_iba     <- plot_oneMarker("BCFA (blood)", "Iso-butyric acid (Blood)")     + theme(axis.text.x  = element_blank())
gg_iva     <- plot_oneMarker("BCFA (blood)", "Iso-valeric acid (Blood)")     + theme(strip.text.x = element_blank())
gg_2ma     <- plot_oneMarker("BCFA (blood)", "2-Methylbutyric acid (Blood)") + theme(strip.text.x = element_blank())

(gg_aca + gg_ppa) / (gg_bta + gg_vla)
(gg_hxa + gg_iba) / (gg_iva + gg_2ma)
```


## Short / Branch chain fatty acids (Stool measurements)

```{r plot stool SCFAs}
#| warning: false
#| message: false
#| fig.width: 10
#| fig.height: 6

gg_aca     <- plot_oneMarker("SCFA (stool)", "Acetic acid (Stool)")          + theme(axis.text.x  = element_blank())
gg_ppa     <- plot_oneMarker("SCFA (stool)", "Propionic acid (Stool)")       + theme(axis.text.x  = element_blank())
gg_bta     <- plot_oneMarker("SCFA (stool)", "Butyric acid (Stool)")         + theme(strip.text.x = element_blank())
gg_vla     <- plot_oneMarker("SCFA (stool)", "Valeric acid (Stool)")         + theme(strip.text.x = element_blank())
gg_hxa     <- plot_oneMarker("SCFA (stool)", "Hexanoic acid (Stool)")        + theme(axis.text.x  = element_blank())

gg_iba     <- plot_oneMarker("BCFA (stool)", "Iso-butyric acid (Stool)")     + theme(axis.text.x  = element_blank())
gg_iva     <- plot_oneMarker("BCFA (stool)", "Iso-valeric acid (Stool)")     + theme(strip.text.x = element_blank())
gg_2ma     <- plot_oneMarker("BCFA (stool)", "2-Methylbutyric acid (Stool)") + theme(strip.text.x = element_blank())

(gg_aca + gg_ppa) / (gg_bta + gg_vla)
(gg_hxa + gg_iba) / (gg_iva + gg_2ma)
```


## SCFA / BCFA blood vs. stool

##### Correlations

Spearman correlations between blood and stool measurements:

```{r}
cor_aca <- cor(dat$blood_aceticAcid,        dat$stool_aceticAcid,        method = "spearman")
cor_ppa <- cor(dat$blood_propionicAcid,     dat$stool_propionicAcid,     method = "spearman")
cor_bta <- cor(dat$blood_butyricAcid,       dat$stool_butyricAcid,       method = "spearman")
cor_vla <- cor(dat$blood_valericAcid,       dat$stool_valericAcid,       method = "spearman")
cor_hxa <- cor(dat$blood_hexanoicAcid,      dat$stool_hexanoicAcid,      method = "spearman")
cor_iba <- cor(dat$blood_isobutyricAcid,    dat$stool_isobutyricAcid,    method = "spearman")
cor_iva <- cor(dat$blood_isovalericAcid,    dat$stool_isovalericAcid,    method = "spearman")
cor_2ma <- cor(dat$blood_methylbutyricAcid, dat$stool_methylbutyricAcid, method = "spearman")
```

- Acetic acid: `r round(cor_aca, 2)`
- Propionic acid: `r round(cor_ppa, 2)`
- Butyric acid: `r round(cor_bta, 2)`
- Valeric acid: `r round(cor_vla, 2)`
- Hexanoic acid: `r round(cor_hxa, 2)`
- Isobutyric acid: `r round(cor_iba, 2)`
- Isovaleric acid: `r round(cor_iva, 2)`
- Methylbutyric acid: `r round(cor_2ma, 2)`


##### Figure for supplements

```{r}
# helper function
plot_bloodVSstoolSCFA <- function(dat, scfa_marker, log_stool = FALSE, log_blood = FALSE) {
  
  colnames(dat)[colnames(dat) == paste0("stool_",scfa_marker)] <- "x_stool"
  colnames(dat)[colnames(dat) == paste0("blood_",scfa_marker)] <- "x_blood"
  
  if (log_stool) {
    dat$x_stool <- log(dat$x_stool + 1)
  }
  if (log_blood) {
    dat$x_blood <- log(dat$x_blood + 1)
  }
  
  # smooth_color <- ifelse(paste0("blood_",scfa_marker) %in% SCFA_bloodMarkers,
  #                        col_vector2["scfa (blood)"], col_vector2["bcfa (blood)"])
  
  x_label <- ifelse(paste0("blood_",scfa_marker) %in% SCFA_bloodMarkers,
                    names(SCFA_bloodMarkers)[match(paste0("blood_",scfa_marker), SCFA_bloodMarkers)],
                    names(BCFA_bloodMarkers)[match(paste0("blood_",scfa_marker), BCFA_bloodMarkers)]) %>% 
    gsub(pattern = " \\(Blood)", replacement = "")
  x_bloodUnit <- marker_units[match(paste0(x_label, " (Blood)"), names(marker_units))] %>% unname()
  x_stoolUnit <- marker_units[match(paste0(x_label, " (Stool)"), names(marker_units))] %>% unname()
  
  ggplot(dat, aes(x = x_blood, y = x_stool)) +
    geom_point(col = "gray30", size = 3, stroke = 0, alpha = 0.6) +
    geom_smooth(se = FALSE, col = "#009ACD", lwd = 2, span = 1) +
    ggtitle(x_label) +
    xlab(paste0(ifelse(log_blood, "Log. blood", "Blood")," measurements [",x_bloodUnit,"]")) +
    ylab(paste0(ifelse(log_stool, "Log. stool", "Stool")," measurements [",x_stoolUnit,"]")) +
    theme_minimal(base_size = 16) +
    theme(plot.title = element_text(hjust = 0.5))
}
```

```{r}
#| warning: false
#| message: false
#| fig.width: 20
#| fig.height: 18

gg_aca    <- plot_bloodVSstoolSCFA(dat, "aceticAcid",    log_blood = TRUE) + theme(panel.grid.minor = element_blank())
gg_ppa    <- plot_bloodVSstoolSCFA(dat, "propionicAcid", log_blood = TRUE) + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())
gg_bta    <- plot_bloodVSstoolSCFA(dat, "butyricAcid",   log_blood = TRUE) + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())
gg_vla    <- plot_bloodVSstoolSCFA(dat, "valericAcid",   log_blood = TRUE) + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())
gg_hxa    <- plot_bloodVSstoolSCFA(dat, "hexanoicAcid") + theme(panel.grid.minor = element_blank())
gg_iba    <- plot_bloodVSstoolSCFA(dat, "isobutyricAcid") + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())
gg_iva    <- plot_bloodVSstoolSCFA(dat, "isovalericAcid", log_blood = TRUE) + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())
gg_2ma    <- plot_bloodVSstoolSCFA(dat, "methylbutyricAcid") + theme(axis.title.y = element_blank(), panel.grid.minor = element_blank())

(gg_aca | gg_ppa | gg_bta | gg_vla) /
  (gg_hxa | gg_iba | gg_iva | gg_2ma)
# ggsave("FigureS15_SCFAs_stoolVSblood.png", width = 16, height = 6)
```



# Further joint version of figures

```{r}
#| column: page
#| fig-width: 16
#| fig-height: 15
#| warning: false
#| message: false

gg_frc     <- plot_oneMarker("Metabolites", "Fructosamine") + theme(axis.text.x  = element_blank())
gg_glc     <- plot_oneMarker("Metabolites", "Glucose")      + theme(axis.text.x  = element_blank())
gg_ins     <- plot_oneMarker("Metabolites", "Insulin")      + theme(axis.text.x = element_blank())
gg_il6     <- plot_oneMarker("Inflammation", "IL-6",    y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_hscrp   <- plot_oneMarker("Inflammation", "hsCRP",   y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_tnfr2   <- plot_oneMarker("Inflammation", "sTNFR-2", y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_rage    <- plot_oneMarker("Inflammation", "sRAGE",   y_log10 = TRUE) + theme(axis.text.x = element_blank())
gg_lbp     <- plot_oneMarker("Leaky gut", "LBP")     + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_fabp2   <- plot_oneMarker("Leaky gut", "FABP-2")  + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_zonulin <- plot_oneMarker("Leaky gut", "Zonulin") + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_aca     <- plot_oneMarker("SCFA (blood)", "Acetic acid (Blood)")    + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_ppa     <- plot_oneMarker("SCFA (blood)", "Propionic acid (Blood)") + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_bta     <- plot_oneMarker("SCFA (blood)", "Butyric acid (Blood)")   + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_vla     <- plot_oneMarker("SCFA (blood)", "Valeric acid (Blood)")   + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_hxa     <- plot_oneMarker("SCFA (blood)", "Hexanoic acid (Blood)")        + theme(strip.text.x = element_blank())
gg_iba     <- plot_oneMarker("BCFA (blood)", "Iso-butyric acid (Blood)")     + theme(strip.text.x = element_blank())
gg_iva     <- plot_oneMarker("BCFA (blood)", "Iso-valeric acid (Blood)")     + theme(strip.text.x = element_blank())
gg_2ma     <- plot_oneMarker("BCFA (blood)", "2-Methylbutyric acid (Blood)") + theme(strip.text.x = element_blank())


row1 <- (gg_frc + gg_glc   + gg_ins     + ggplot()) + patchwork::plot_layout(nrow = 1, widths = c(1.01,.98,1.03,1.18))
row2 <- (gg_il6 + gg_hscrp + gg_tnfr2   + gg_rage)  + patchwork::plot_layout(nrow = 1, widths = 1)
row3 <- (gg_lbp + gg_fabp2 + gg_zonulin + ggplot()) + patchwork::plot_layout(nrow = 1, widths = c(1.01,.98,1.03,1.18))
row4 <- (gg_aca + gg_ppa   + gg_bta     + gg_vla)   + patchwork::plot_layout(nrow = 1, widths = 1)
row5 <- (gg_hxa + gg_iba   + gg_iva     + gg_2ma)   + patchwork::plot_layout(nrow = 1, widths = 1)

row1 / row2 / row3 / row4 / row5
# ggsave("descriptives_allMarkers.png", width = 16, height = 12.5)
```


```{r}
#| column: page
#| fig-width: 16
#| fig-height: 6
#| warning: false
#| message: false

gg_il6     <- plot_oneMarker("Inflammation", "IL-6",   y_log10 = TRUE) + theme(axis.text.x = element_blank())
gg_hscrp   <- plot_oneMarker("Inflammation", "hsCRP", y_log10 = TRUE) + theme(axis.text.x = element_blank())
gg_tnfr2   <- plot_oneMarker("Inflammation", "sTNFR-2", y_log10 = TRUE) + theme(axis.text.x = element_blank())
gg_rage    <- plot_oneMarker("Inflammation", "sRAGE",  y_log10 = TRUE)
gg_lbp     <- plot_oneMarker("Leaky gut", "LBP")
gg_fabp2   <- plot_oneMarker("Leaky gut", "FABP-2")
gg_zonulin <- plot_oneMarker("Leaky gut", "Zonulin")

row1 <- (gg_il6 + gg_hscrp + gg_tnfr2 + gg_rage) + patchwork::plot_layout(nrow = 1, widths = 1)
row2 <- (gg_lbp + gg_fabp2 + gg_zonulin + ggplot()) + patchwork::plot_layout(nrow = 1, widths = c(1.01,.98,1.03,1.18))

row1 / row2
# ggsave("descriptives_inflam_leaky.png", width = 16, height = 5)
```


```{r}
#| column: page
#| fig-width: 16
#| fig-height: 9
#| warning: false
#| message: false

gg_frc     <- plot_oneMarker("Metabolites", "Fructosamine") + theme(axis.text.x  = element_blank())
gg_glc     <- plot_oneMarker("Metabolites", "Glucose")      + theme(axis.text.x  = element_blank())
gg_ins     <- plot_oneMarker("Metabolites", "Insulin")      + theme(axis.text.x = element_blank())
gg_il6     <- plot_oneMarker("Inflammation", "IL-6",    y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_hscrp   <- plot_oneMarker("Inflammation", "hsCRP",   y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_tnfr2   <- plot_oneMarker("Inflammation", "sTNFR-2", y_log10 = TRUE) + theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_rage    <- plot_oneMarker("Inflammation", "sRAGE",   y_log10 = TRUE) + theme(axis.text.x = element_blank())
gg_lbp     <- plot_oneMarker("Leaky gut", "LBP",     col_vec = col_vector_withoutSCFAs, col_vec2 = col_vector_withoutSCFAs2) +
  theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_fabp2   <- plot_oneMarker("Leaky gut", "FABP-2",  col_vec = col_vector_withoutSCFAs, col_vec2 = col_vector_withoutSCFAs2) +
  theme(axis.text.x = element_blank(), strip.text.x = element_blank())
gg_zonulin <- plot_oneMarker("Leaky gut", "Zonulin", col_vec = col_vector_withoutSCFAs, col_vec2 = col_vector_withoutSCFAs2) +
  theme(axis.text.x = element_blank(), strip.text.x = element_blank())


row1 <- (gg_frc + gg_glc   + gg_ins     + ggplot()) + patchwork::plot_layout(nrow = 1, widths = c(1.01,.98,1.03,1.18))
row2 <- (gg_il6 + gg_hscrp + gg_tnfr2   + gg_rage)  + patchwork::plot_layout(nrow = 1, widths = 1)
row3 <- (gg_lbp + gg_fabp2 + gg_zonulin + ggplot()) + patchwork::plot_layout(nrow = 1, widths = c(1.01,.98,1.03,1.18))

row1 / row2 / row3
# ggsave("descriptives_allMarkers_withoutSCFAs.png", width = 16, height = 7.5)
```
