---
title: "Description Microbiome Changes with focus on phylum level"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Data preparation

```{r packages}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(SummarizedExperiment) # microbiome analysis

library(dplyr)      # data handling
library(tidyr)      # data transformation
library(ggplot2)    # data visualization
library(kableExtra) # table printing

theme_set(theme_minimal()) # set ggplot2 theme
```

```{r set file paths}
define_dataPaths()
```

```{r data preparation}
#| warning: false

tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")

dat <- tse %>% 
  colData() %>% 
  as.data.frame()
```

## Choose three random participants

... for which to create some the following visualizations

```{r participant choice}
participants <- dat$participant_id %>% unique() %>% as.character() %>% sample(size = 3) %>% sort()
```

```{r participant data preparation}
datShort_list <- lapply(participants, function(x) {
  
  # retrieve the sample names for the participant
  dat_sampleIDs <- dat %>% 
    filter(participant_id == x) %>% 
    select(inter_treat)
  ids_samples        <- row.names(dat_sampleIDs)
  names(ids_samples) <- dat_sampleIDs$inter_treat
  
  # create a dataset with all taxonomic groupings in columns
  dat_tax <- tse %>% 
    rowData() %>% 
    as.data.frame() %>% 
    mutate(species = row.names(.))
  row.names(dat_tax) <- NULL
  
  # create one base dataset for all figures
  dat_relAbd <- tse %>% 
    assay("relAbd") %>% 
    as.data.frame() %>% 
    select(all_of(ids_samples)) %>% 
    mutate(species = row.names(.)) %>% 
    pivot_longer(cols = -species, names_to = "inter_treat", values_to = "relative_abundance") %>% 
    select(inter_treat, relative_abundance, species) %>% 
    mutate(inter_treat = factor(inter_treat, levels = c("Baseline.Fresh","After.Fresh","Baseline.Pasteurized",
                                                        "After.Pasteurized","Baseline.FollowUp"))) %>% 
    dplyr::left_join(dat_tax, by = "species")
  
})
```


# Changes on phylum level -- overall

Let's analyze the changes of the five overall most abundant phyla.

```{r general data preparation}
# create a dataset with all taxonomic groupings in columns
dat_tax <- tse %>% 
  rowData() %>% 
  as.data.frame() %>% 
  mutate(species = row.names(.))
row.names(dat_tax) <- NULL

# create a dataset for matching the intervention and participant id variables to the samples
dat_samples <- dat %>% 
  select(participant_id, inter_treat) %>% 
  mutate(sample = row.names(.))
row.names(dat_samples) <- NULL

# create one base dataset for all figures
dat_relAbd <- tse %>% 
  assay("relAbd") %>% 
  as.data.frame() %>% 
  mutate(species = row.names(.)) %>% 
  pivot_longer(cols = -species, names_to = "sample", values_to = "relative_abundance") %>% 
  dplyr::left_join(dat_samples, by = "sample") %>% 
  mutate(inter_treat  = factor(inter_treat, levels = c("Baseline.Fresh","After.Fresh","Baseline.Pasteurized",
                                                       "After.Pasteurized","Baseline.FollowUp")),
         treatment    = case_when(grepl("Baseline", inter_treat) ~ "Baseline",
                                  grepl("After",    inter_treat) ~ "After",
                                  TRUE                           ~ NA_character_),
         treatment    = factor(treatment, levels = c("Baseline", "After")),
         intervention = case_when(grepl("Fresh",       inter_treat) ~ "Fresh",
                                  grepl("Pasteurized", inter_treat) ~ "Pasteurized",
                                  grepl("FollowUp",    inter_treat) ~ "FollowUp",
                                  TRUE                              ~ NA_character_),
         intervention = factor(intervention, levels = c("Fresh", "Pasteurized", "FollowUp"))) %>% 
  select(participant_id, inter_treat, treatment, intervention, relative_abundance, species) %>% 
  dplyr::left_join(dat_tax, by = "species")
```

```{r create a table}
#| message: false

# extract the five most common phyla
phyla_vec <- dat_relAbd %>% 
  group_by(phylum) %>% 
  summarize(sum_relAbd = sum(relative_abundance, na.rm = TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(sum_relAbd)) %>% 
  dplyr::slice(1:5) %>% 
  pull(phylum) %>% 
  as.character()

# restrict the data to these phyla, aggregate it and remove the follow-up measurements
dat_relAbd_phylum <- dat_relAbd %>% 
  filter(phylum %in% phyla_vec) %>% 
  group_by(participant_id, inter_treat, phylum) %>% 
  summarize(relative_abundance = sum(relative_abundance, na.rm = TRUE),
            intervention       = first(intervention),
            treatment          = first(treatment)) %>% 
  ungroup() %>% 
  filter(intervention != "FollowUp")

# create plots and tables
results_list <- lapply(phyla_vec, function(x) {
  dat_relAbd_phylum %>% 
    filter(phylum == x) %>% 
    plot_crossoverBoxplots(y_var            = "relative_abundance",
                           intervention_var = "intervention",
                           t_var            = "treatment",
                           subject_var      = "participant_id",
                           title            = paste("Phylum", x))
})

# print plots and tables
results_list[[1]]$ggplot_object
results_list[[1]]$results_table %>% kable() %>% kable_styling()
results_list[[2]]$ggplot_object
results_list[[2]]$results_table %>% kable() %>% kable_styling()
results_list[[3]]$ggplot_object
results_list[[3]]$results_table %>% kable() %>% kable_styling()
results_list[[4]]$ggplot_object
results_list[[4]]$results_table %>% kable() %>% kable_styling()
results_list[[5]]$ggplot_object
results_list[[5]]$results_table %>% kable() %>% kable_styling()
```


# Changes on phylum level -- exemplaric participants 

Each graphic only shows the top five most abundant bacteria species (reference:
sum over all time points) and puts all other categories in the "Others" group.

```{r general plots}
#| message: false
#| fig.width: 10

# final preparation for the plot data
plotDat_list <- lapply(seq_len(length(participants)), function(i) {
  
  dat_short <- datShort_list[[i]]
  
  # categorize less abundant bacteria groups into "Others"
  tab_phylum <- dat_short %>% 
    group_by(phylum) %>% 
    summarize(sum_relAbd = sum(relative_abundance)) %>% 
    ungroup() %>% 
    arrange(desc(sum_relAbd))
  
  # aggregate the relative abundances per bacteria group
  dat_short %>% 
    mutate(phylum_cat = case_when(phylum %in% tab_phylum$phylum[1:5] ~ phylum,
                                  TRUE                               ~ "Others")) %>% 
    ungroup() %>% 
    group_by(inter_treat, phylum_cat) %>% 
    summarize(relative_abundance = sum(relative_abundance, na.rm = TRUE)) %>% 
    ungroup() %>% 
    mutate(phylum_cat = factor(phylum_cat, levels = rev(c(as.character(tab_phylum$phylum[1:5]), "Others"))))
})


# prepare plots
gg_list <- lapply(seq_len(length(participants)), function(i) {
  
  dat_plot    <- plotDat_list[[i]]
  participant <- participants[i]
  
  dat_plot %>% 
    ggplot(aes(x = inter_treat, weight = relative_abundance, fill = phylum_cat)) +
    geom_bar() +
    geom_hline(yintercept = c(25,50,75), linetype = 2, col = "gray30") +
    ggtitle("Verteilung der Bakterien nach Stämmen", subtitle = paste("Proband", participant)) +
    scale_y_continuous(labels = function(x) { paste0(x, "%") }) +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5),
          axis.title    = element_blank(),
          legend.title  = element_blank(),
          axis.text.x   = element_text(angle = 45, hjust = 1))
  
})


# prepare tables
tab_list <- lapply(seq_len(length(participants)), function(i) {
  
  plotDat_list[[i]] %>%
    pivot_wider(values_from = "relative_abundance", id_cols = "phylum_cat", names_from = "inter_treat") %>% 
    kable() %>% 
    kable_styling()
  
})

gg_list[[1]]
tab_list[[1]]
gg_list[[2]]
tab_list[[2]]
gg_list[[3]]
tab_list[[3]]
```
