---
title: "Beta diversity -- unweighted UniFrac distance"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Data preparation

```{r packages}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(SummarizedExperiment) # microbiome analysis
library(vegan)                # distance metric calculation

library(dplyr)      # data handling
library(tidyr)      # data transformation
library(ggplot2)    # data visualization
library(patchwork)  # aligning ggplot objects in a grid
library(kableExtra) # table printing

# set ggplot2 theme
theme_set(
  theme_minimal() +
    theme(plot.title    = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5))
)
```

```{r set file paths}
define_dataPaths()
```

```{r data preparation}
#| warning: false

tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")
```


# Calculate unweighted UniFrac distance matrix

```{r}
#| warning: false

datDist_list <- calculate_distances(tse, "relAbd")

datDist_intra <- datDist_list$datDist_intraPersonal
datDist_inter <- datDist_list$datDist_interPersonal
```

```{r}
# add strata information
datDist_intra <- datDist_intra %>% 
  mutate(age_cat        = case_when(age < 50 ~ paste0(min(age), "-49"),
                                    TRUE     ~ paste0("50-", max(age))),
         age_cat        = factor(age_cat),
         bmi_cat        = case_when(bmi_t0 < 25 ~ paste0("[", floor(min(bmi_t0)), ", 25)"),
                                    TRUE        ~ paste0("[25, ", ceiling(max(bmi_t0)), "]")),
         bmi_cat        = factor(bmi_cat))
```

The pairwise unweighted UniFrac distances in our sample vary between
`r round(min(datDist_intra$uniFrac_unweighted), 2)` and `r round(max(datDist_intra$uniFrac_unweighted), 2)`
regarding the intra-personal variability and between
`r round(min(datDist_inter$uniFrac_unweighted), 2)` and `r round(max(datDist_inter$uniFrac_unweighted), 2)`
regarding the inter-personal variability.
The unweighted UniFrac distances have the following distribution:

```{r}
#| message: false
#| warning: false

gg1 <- ggplot(datDist_intra, aes(x = uniFrac_unweighted)) + geom_histogram() + xlim(c(0,1)) + ggtitle("Intra-personal unweighted UniFrac distances")
gg2 <- ggplot(datDist_inter, aes(x = uniFrac_unweighted)) + geom_histogram() + xlim(c(0,1)) + ggtitle("Inter-personal unweighted UniFrac distances")

gg1 / gg2
```


# Intra-personal variation

1. For each timepoint comparison (e.g. *T1 vs. T3*) one unweighted UniFrac distance was
calculated per person
2. The following boxplots all depict such comparisons where one timepoint
measurement of a person was compared to another timepoint measurement of the same
person. Accordingly, each point encodes **intra-personal variation**.
3. The displayed dissimilarities vary between 0 (*perfect similarity*)
and 1 (*no similarity at all*).


## Development over the timepoints

```{r}
datDist_intra %>% 
  filter(comparison %in% c("T1 vs. T2", "T2 vs. T3", "T3 vs. T4", "T4 vs. T5")) %>% 
  ggplot(aes(x = comparison, y = uniFrac_unweighted)) +
  geom_boxplot() +
  geom_line(aes(group = participant_id, col = participant_id), alpha = 0.3) +
  facet_wrap(~ group) +
  theme(axis.title.x    = element_blank(),
        legend.position = "none")
```


## Variation by intervention

```{r}
#| message: false

datDist_intra %>% 
  filter(comparison_group != "other") %>% 
  ggplot(aes(x = comparison, y = uniFrac_unweighted)) +
  geom_boxplot() +
  geom_line(aes(group = participant_id, col = participant_id), alpha = 0.3) +
  facet_wrap(~ comparison_group, scales = "free_x") +
  theme(axis.title.x    = element_blank(),
        legend.position = "none")
```


# Inter-personal variation

```{r}
datDist_inter %>% 
  filter(timepoint_group == "baselines") %>% 
  ggplot(aes(x = timepoint, y = uniFrac_unweighted)) +
  geom_boxplot() +
  ylim(c(0,1)) +
  theme(axis.title.x = element_blank())
```


# Stratified analysis by age, gender, BMI

```{r}
#| column: page
#| fig-width: 12

# by age
datDist_intra %>% 
  filter(comparison_group != "other",
         comparison != "T1 vs. T5") %>% 
  mutate(comparison_group = case_when(comparison_group == "baselines" & comparison == "T1 vs. T3" ~ "Baseline variation\n(1st vs. 2nd baseline)",
                                      comparison_group == "baselines" & comparison == "T3 vs. T5" ~ "Follow-up variation\n(follow-up vs. 2nd baseline)",
                                      comparison_group == "fresh intervention"                    ~ "Fresh intervention\n(vs. baseline)",
                                      comparison_group == "pasteurized intervention"              ~ "Pasteurized intervention\n(vs. baseline)"),
         comparison_group = factor(comparison_group,
                                   levels = c("Baseline variation\n(1st vs. 2nd baseline)", "Fresh intervention\n(vs. baseline)",
                                              "Pasteurized intervention\n(vs. baseline)", "Follow-up variation\n(follow-up vs. 2nd baseline)"))) %>% 
  ggplot(aes(x = comparison_group, y = uniFrac_unweighted, col = comparison_group)) +
  geom_boxplot() +
  scale_y_continuous("Unweighted Unifrac dissimilarity") +
  scale_color_manual(values = c("gray20","#1F78B4","#FF7F00","gray20")) +
  facet_wrap(~ age_cat) +
  ggtitle("Beta diversity", "(baseline and intervention)") +
  theme(axis.title.x    = element_blank(),
        legend.position = "none",
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray95"))


# by gender
datDist_intra %>% 
  filter(comparison_group != "other",
         comparison != "T1 vs. T5") %>% 
  mutate(comparison_group = case_when(comparison_group == "baselines" & comparison == "T1 vs. T3" ~ "Baseline variation\n(1st vs. 2nd baseline)",
                                      comparison_group == "baselines" & comparison == "T3 vs. T5" ~ "Follow-up variation\n(follow-up vs. 2nd baseline)",
                                      comparison_group == "fresh intervention"                    ~ "Fresh intervention\n(vs. baseline)",
                                      comparison_group == "pasteurized intervention"              ~ "Pasteurized intervention\n(vs. baseline)"),
         comparison_group = factor(comparison_group,
                                   levels = c("Baseline variation\n(1st vs. 2nd baseline)", "Fresh intervention\n(vs. baseline)",
                                              "Pasteurized intervention\n(vs. baseline)", "Follow-up variation\n(follow-up vs. 2nd baseline)"))) %>% 
  ggplot(aes(x = comparison_group, y = uniFrac_unweighted, col = comparison_group)) +
  geom_boxplot() +
  scale_y_continuous("Unweighted Unifrac dissimilarity") +
  scale_color_manual(values = c("gray20","#1F78B4","#FF7F00","gray20")) +
  facet_wrap(~ gender) +
  ggtitle("Beta diversity", "(baseline and intervention)") +
  theme(axis.title.x    = element_blank(),
        legend.position = "none",
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray95"))



# by BMI
datDist_intra %>% 
  filter(comparison_group != "other",
         comparison != "T1 vs. T5") %>% 
  mutate(comparison_group = case_when(comparison_group == "baselines" & comparison == "T1 vs. T3" ~ "Baseline variation\n(1st vs. 2nd baseline)",
                                      comparison_group == "baselines" & comparison == "T3 vs. T5" ~ "Follow-up variation\n(follow-up vs. 2nd baseline)",
                                      comparison_group == "fresh intervention"                    ~ "Fresh intervention\n(vs. baseline)",
                                      comparison_group == "pasteurized intervention"              ~ "Pasteurized intervention\n(vs. baseline)"),
         comparison_group = factor(comparison_group,
                                   levels = c("Baseline variation\n(1st vs. 2nd baseline)", "Fresh intervention\n(vs. baseline)",
                                              "Pasteurized intervention\n(vs. baseline)", "Follow-up variation\n(follow-up vs. 2nd baseline)"))) %>% 
  ggplot(aes(x = comparison_group, y = uniFrac_unweighted, col = comparison_group)) +
  geom_boxplot() +
  scale_y_continuous("Unweighted Unifrac dissimilarity") +
  scale_color_manual(values = c("gray20","#1F78B4","#FF7F00","gray20")) +
  facet_wrap(~ bmi_cat) +
  ggtitle("Beta diversity", "(baseline and intervention)") +
  theme(axis.title.x    = element_blank(),
        legend.position = "none",
        panel.grid.minor   = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray95"))
```


# PCoA on unweighted UniFrac distances (all comparisons)

```{r}
# calculate the unweighted UniFrac distance matrix
distMatrix_uUF <- tse %>% mia::calculateUnifrac(assay.type = "relAbd", weighted = FALSE)

# PCoA
pcoa_uUF <- ecodist::pco(distMatrix_uUF)

# All components could be found here: 
# pcoa_uUF$vectors
# But we only need the first two to demonstrate what we can do:
pcoa_df <- data.frame(pcoa1 = pcoa_uUF$vectors[,1], 
                      pcoa2 = pcoa_uUF$vectors[,2])

# Create a plot
pcoa_df %>% 
  mutate(inter_treat = colData(tse)$inter_treat) %>% 
  ggplot(data = ., aes(x = pcoa1, y = pcoa2, color = inter_treat)) +
  geom_point() +
  stat_ellipse() +
  scale_color_brewer(palette = "Dark2") +
  ggtitle("unweighted UniFrac-based PCoA") +
  xlab("PC1") + ylab("PC2")
```
