---
title: "Dimension Reduction for Taxonomy Data -- Main Results"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

This document contains the results of the application of the UMAP dimension reduction
(also called "ordination") method to the microbiome data.
An overview of all evaluated alternative methods (which lead to similar results
as UMAP, though) can be found in a separate document.

How to read it:

- UMAP performs a **(dis)similarity analysis on relative abundances**.
In the estimation process, every sample is compared to every other sample in the
data based on the similarity of their bacteria species profile. **The more
similar two samples are, the closer their points are plotted in the different graphs.**
- Note: For UMAP as a nonlinear method there is no nontrivial measure to
estimate how much information is lost by focusing on two dimensions only.
For the PCoA methods the first two dimensions only explain 15% of the overall
variation in the data, i.e. **85% of the information is lost when focusing only
on these two dimensions**. Accordingly, **results should be interpreted with
care**.
- As of this, minor differences in the plotted ellipses should not be overinterpreted,
but instead **interpretation should focus on more global patterns** like neighboring
families / genuses or the visualized age and richness patterns plotted for the
UMAP results.


## Data preparation and UMAP estimation

```{r packages}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(SummarizedExperiment) # microbiome analysis
library(mia)                  # microbiome analysis
library(vegan)                # (dis)similarity measures

library(scater)     # dimension reduction and visualizations

library(dplyr)      # data handling
library(tidyr)      # data transformation
library(ggplot2)    # data visualization
library(ggpubr)     # joint ggplots
library(patchwork)  # joint ggplots
library(kableExtra) # table printing

# set ggplot2 theme
theme_set(
  theme_minimal() +
    theme(plot.title       = element_text(hjust = 0.5),
          plot.subtitle    = element_text(hjust = 0.5),
          panel.grid.minor = element_blank(),
          plot.background  = element_rect(fill = "white", color = "white"))
)
```

```{r set file paths}
define_dataPaths()
```

```{r data preparation}
#| warning: false

tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")

dat <- tse %>% 
  colData() %>% 
  as.data.frame()
```

```{r}
# add logarithmized versions of the SCFA variables to the dataset, for plotting
dat <- dat %>% 
  mutate(stool_methylbutyricAcid_log = log(stool_methylbutyricAcid),
         stool_aceticAcid_log         = log(stool_aceticAcid),
         stool_butyricAcid_log        = log(stool_butyricAcid),
         stool_hexanoicAcid_log       = log(stool_hexanoicAcid),
         stool_isobutyricAcid_log     = log(stool_isobutyricAcid),
         stool_isovalericAcid_log     = log(stool_isovalericAcid),
         stool_propionicAcid_log      = log(stool_propionicAcid),
         stool_valericAcid_log        = log(stool_valericAcid))

colData(tse) <- dat %>% DataFrame()
```


```{r estimate UMAP}
#| warning: false
#| message: false
#| results: hide

set.seed(2025)
tse <- tse %>% runUMAP(name = "UMAP", exprs_values = "relAbd")
```

```{r logarithmize some variables, for plotting}
colData(tse)$logDailyConsumption_fiber <- log(colData(tse)$dailyConsumption_fiber)
colData(tse)$logDailyConsumption_fat   <- log(colData(tse)$dailyConsumption_fat)
colData(tse)$logDailyConsumption_water <- log(colData(tse)$dailyConsumption_water)
```


```{r}
# helper function
get_plotList <- function(taxRank_dominance, taxRank_topNfocus) {
  
  # create vector of all biodiversity measures in the dataset
  covars_bd <- dat %>% 
    select(starts_with("richness") | starts_with("evenness") | starts_with("diversity") |
             starts_with("dominance") | starts_with("rarity") | starts_with("divergence"),
           stool_type, ph_value, contains("Acid")) %>% 
    colnames()
  
  # get plot list
  plot_dimReduction(tse                = tse,
                    dimReduction_name  = "UMAP",
                    assay              = "relAbd",
                    taxRank_dominance  = taxRank_dominance,
                    taxRank_topNfocus  = taxRank_topNfocus,
                    covars_categorical = c("gender", "flatulence_afterFresh", "flatulence_afterPast",
                                           "betterDigestion_afterFresh", "betterDigestion_afterPast",
                                           "stoolType_healthy", "diet_veggie"),
                    covars_metric      = c("age", "bmi_t0", "dailyConsumption_fiber",
                                           "dailyConsumption_fat", "dailyConsumption_water",
                                           "logDailyConsumption_fiber", "logDailyConsumption_fat",
                                           "logDailyConsumption_water",
                                           covars_bd))
}
```

```{r create plot objects}
#| warning: false

gg_famList <- get_plotList("family",  taxRank_topNfocus = 8)
gg_genList <- get_plotList("genus",   taxRank_topNfocus = 10)
gg_spcList <- get_plotList("species", taxRank_topNfocus = 13)
```


# Rel. abundance of L. paracasei

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 16

gg_famList$gg_Lparacasei_byIntervention / gg_famList$gg_Lparacasei_byIntervention_cat

gg_famList$gg_LparacaseiChange_byIntervention / gg_famList$gg_LparacaseiChange_byIntervention_cat
```


# Most dominant taxa

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 16

(gg_famList$gg_dominantTaxa + gg_genList$gg_dominantTaxa) / (gg_spcList$gg_dominantTaxa + guides(color = guide_legend(ncol = 2))) + plot_layout(guides = "collect")
```


## Figure for publication

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 6

gg_spcList$gg_dominantTaxa +
  ggtitle("UMAP visualization of stool microbiota",
          "colored by dominant species") +
  theme(panel.grid.major = element_line(color = "gray95"))
# ggsave("UMAP_dominantSpecies.png", width = 18, height = 7)
```


# Strongest growth taxa

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 8

((gg_famList$gg_maxGrowthTaxa_fresh + theme(legend.position = "none")) + gg_famList$gg_maxGrowthTaxa_past) + plot_layout(guides = "collect")
((gg_genList$gg_maxGrowthTaxa_fresh + theme(legend.position = "none")) + gg_genList$gg_maxGrowthTaxa_past) + plot_layout(guides = "collect")
((gg_spcList$gg_maxGrowthTaxa_fresh + theme(legend.position = "none")) + gg_spcList$gg_maxGrowthTaxa_past) + plot_layout(guides = "collect")
```


# Strongest loss taxa

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 8

((gg_famList$gg_maxLossTaxa_fresh + theme(legend.position = "none")) + gg_famList$gg_maxLossTaxa_past) + plot_layout(guides = "collect")
((gg_genList$gg_maxLossTaxa_fresh + theme(legend.position = "none")) + gg_genList$gg_maxLossTaxa_past) + plot_layout(guides = "collect")
((gg_spcList$gg_maxLossTaxa_fresh + theme(legend.position = "none")) + gg_spcList$gg_maxLossTaxa_past) + plot_layout(guides = "collect")
```


# Experimental settings

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 8

(gg_famList$gg_groups + gg_famList$gg_interventionTimes) + plot_layout(guides = "collect")
```


## Figure for publication

```{r}
gg_famList$gg_interventionTimes_v2 +
  ggtitle("UMAP visualization of stool microbiota",
          "colored by intervention") +
  theme(panel.grid.major = element_line(color = "gray95"))
# ggsave("UMAP_interventionEffects.png", width = 8, height = 6)
```


# Variation of each participants' measurements

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 12

# plot all data
gg_famList$gg_participantVariation

# ... without T5
gg_famList$gg_participantVariation_withoutT5
```


```{r}
#| column: page
#| fig.width: 16
#| fig.height: 8

# plot by intervention
gg_famList$gg_participantVariation_byIntervention

# plot by diversity
gg_famList$gg_participantVariation_byInt_div

# plot by high vs. low change in observed richness
gg_famList$gg_participantVariation_byRichness

# plot by high vs. low change in Shannon diversity
gg_famList$gg_participantVariation_byDiversity
```


## Figure for publication

```{r}
#| fig-width: 6
#| fig-height: 6

gg_famList$gg_participantVariation_minimal
# ggsave("FigureS7_UMAP_personalVariation.png", width = 6, height = 6)
```


# Sociodemographics

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 16

gg_famList$covar_gender / (gg_famList$covar_age + gg_famList$covar_bmi_t0) + plot_layout(guides = "collect")
```


# Dietary variables

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 16

(gg_famList$covar_diet_veggie + gg_famList$covar_dailyConsumption_fiber) / (gg_famList$covar_dailyConsumption_fat + gg_famList$covar_dailyConsumption_water) + plot_layout(guides = "collect")

(ggplot() + gg_famList$covar_logDailyConsumption_fiber) / (gg_famList$covar_logDailyConsumption_fat + gg_famList$covar_logDailyConsumption_water) + plot_layout(guides = "collect")
```


# Measures of biodiversity

```{r}
#| column: page
#| fig.width: 16
#| fig.height: 8

# richness
(gg_famList$covar_richness_hill + gg_famList$covar_richness_observed) + plot_layout(guides = "collect")

# evenness
(gg_famList$covar_evenness_pielou + gg_famList$covar_evenness_simpson) + plot_layout(guides = "collect")

# alpha diversity
(gg_famList$covar_diversity_shannon + gg_famList$covar_diversity_invSimpson) + plot_layout(guides = "collect")

# dominance
(gg_famList$covar_dominance_dbp + gg_famList$covar_dominance_coreAbundance) + plot_layout(guides = "collect")

# rarity and divergence
(gg_famList$covar_rarity_logModuloSkewness + gg_famList$covar_divergence_toMedian) + plot_layout(guides = "collect")
```


## Figure for publication

```{r}
#| message: false
#| column: page
#| fig-width: 22
#| fig-height: 13

figS6_baseSize <- 16

# ensure identical plot limits for each of the jointly plotted figures
ylim <- c(-3.5, 3.2)

gg_domTaxa <- gg_spcList$gg_dominantTaxa +
  ggtitle("dominant species", "") +
  theme_minimal(base_size = figS6_baseSize) +
  theme(plot.title       = element_text(hjust = 0.5),
        panel.grid.major = element_line(color = "gray95"),
        panel.grid.minor = element_blank())
gg_domTaxa_legend <- ggpubr::get_legend(gg_domTaxa)
gg_domTaxa        <- gg_domTaxa + theme(legend.position = "none")

gg_gender <- gg_famList$covar_sex +
  ylim(ylim) +
  ggtitle("sex", NULL) +
  theme_minimal(base_size = figS6_baseSize) +
  theme(plot.title      = element_text(hjust = 0.5),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

gg_age <- gg_famList$covar_age +
  ylim(ylim) +
  ggtitle("age", NULL) +
  theme_minimal(base_size = figS6_baseSize) +
  theme(plot.title       = element_text(hjust = 0.5),
        legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        axis.title.y     = element_blank(),
        axis.text.y      = element_blank())

gg_bmi <- gg_famList$covar_bmi_t0 +
  ylim(ylim) +
  ggtitle("baseline BMI", NULL) +
  scale_color_viridis_c("baseline BMI") +
  theme_minimal(base_size = figS6_baseSize) +
  theme(plot.title       = element_text(hjust = 0.5),
        legend.position  = "bottom",
        panel.grid.minor = element_blank(),
        axis.title.y     = element_blank(),
        axis.text.y      = element_blank())

gg_divShannon <- gg_famList$covar_diversity_shannon +
  ggtitle("Shannon diversity", NULL) +
  scale_color_continuous("Shannon diversity", type = "viridis") +
  ylim(ylim) +
  theme_minimal(base_size = figS6_baseSize) +
  theme(plot.title       = element_text(hjust = 0.5),
        panel.grid.major = element_line(color = "gray95"),
        panel.grid.minor = element_blank(),
        legend.position  = "bottom",
        axis.title.y     = element_blank(),
        axis.text.y      = element_blank())


# joint plot
layout <- "
AAABBBBB
CCDDEEFF
"
gg_domTaxa + gg_domTaxa_legend + gg_gender + gg_age + gg_bmi + gg_divShannon +
  patchwork::plot_layout(design = layout, guides = "keep") +
  patchwork::plot_annotation(title = "UMAP colored by ...", theme = theme(plot.title = element_text(hjust = 0.5, size = 20)))
# ggsave("FigureS5_UMAP.png", width = 22, height = 13)
```



# Stool and digestion

In the following 'flatulence after fresh / pasteurized intervention' and 'better digestion after fresh / pasteurized intervenion' plots all measurements of a person are plotted and based on the specific variable (e.g. if a person had flatulence after the fresh intervention) **all timepoints** of this person are colored accordingly.

```{r}
#| warning: false
#| column: page
#| fig.width: 16
#| fig.height: 8

# flatulence after interventions
gg_famList$covar_flatulence_afterFresh + gg_famList$covar_flatulence_afterPast

# better digestion after intervenions
gg_famList$covar_betterDigestion_afterFresh + gg_famList$covar_betterDigestion_afterPast

# stool type
gg_famList$covar_stool_type + gg_famList$covar_stoolType_healthy
```


```{r}
#| fig.width: 8
#| fig.height: 8

# PH value
gg_famList$covar_ph_value
```
