---
title: "Sauerkraut study - Model-based Stool SCFA Analysis"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

```{r packages}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(SummarizedExperiment) # microbiome analysis
library(mia)                  # microbiome analysis
library(mgcv)                 # mixed model estimation
library(mgcViz)               # mgcv-based visualizations
library(APCtools)             # mgcv-based tables and visualizations

library(dplyr)      # data handling
library(tidyr)      # data transformation
library(ggplot2)    # data visualization
library(patchwork)  # grid of ggplots
library(kableExtra) # print tables

# set ggplot2 theme
theme_set(
  theme_minimal() +
    theme(plot.title       = element_text(hjust = 0.5),
          plot.subtitle    = element_text(hjust = 0.5),
          panel.grid.minor = element_blank(),
          plot.background  = element_rect(fill = "white", color = "white"))
)
```

# Methodology

For each stool parameter, the below model is estimated in three variants, as described in the paper:

- Gaussian regression model
- Gaussian regression model with log transformation
- Gamma regression model
- sidenote: additionally, Gaussian with square root transformation was evaluated, but due to only small benefits regarding the models' goodness of fit, and the unintuitive interpretation we'd get from such a square root transformation, it was dropped

Each of the above variants were estimated and the final model was chosen as the model with the highest share of explained deviance per stool parameter.

For each parameter of interest, a separate linear mixed regression model is
estimated.


##### General notes

- follow-up measurements were excluded before model estimation
- internal note: nonlinear effects of age and BMI were evaluated for a lot of models, but did not lead to (structural) improvements of any of the models' predictive power. In most models, it didn't change the explained deviance share at all.
- internal note: age-specific intervention effects were only evaluated for two age groups. Looking at a third group, e.g. at 'below 30', suffers of a quite limited sample size


# Data preparation

```{r set file paths}
define_dataPaths()
```

```{r read data}
#| warning: false

tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")

dat <- tse %>% 
  colData() %>% 
  as.data.frame()
```

```{r compute experiment variables}
# remove follow-up measurements
dat_model <- dat %>% 
  filter(intervention != "FollowUp") %>% 
  mutate(timepoint    = droplevels(timepoint),
         period       = droplevels(period),
         intervention = droplevels(intervention))

# construct the intervention variable and the carry-over variable
dat_model <- dat_model %>% 
  mutate(intervention = case_when(treatment == "Baseline" ~ "Baseline",
                                  TRUE                    ~ intervention),
         carry_over   = case_when(timepoint %in% c("T1", "T2") ~ "not possible (T1, T2)",
                                  group == "Fresh first"       ~ "from fresh intervention",
                                  group == "Pasteurized first" ~ "from pasteurized intervention"),
         carry_over   = factor(carry_over, levels = c("not possible (T1, T2)", "from fresh intervention", "from pasteurized intervention")),
         phase        = case_when(timepoint %in% c("T3","T4") ~ "phase 2",
                                  TRUE                        ~ "phase 1"),
         phase        = factor(phase))

# categorize the age and bmi variables for testing category-specific intervention effects
dat_model <- dat_model %>% 
  mutate(age_centered50 = age - 50,
         bmi_centered25 = bmi_t0 - 25,
         age_cat = case_when(age < 50 ~ paste0(min(age), "-49"),
                             TRUE     ~ paste0("50-", max(age))),
         age_cat = factor(age_cat),
         bmi_cat = case_when(bmi_t0 < 25 ~ paste0("[", floor(min(bmi_t0)), ", 25)"),
                             TRUE        ~ paste0("[25, ", ceiling(max(bmi_t0)), "]")),
         bmi_cat = factor(bmi_cat))
```

```{r lookup vectors}
SCFA_markers         <- c("Acetic acid (Stool)"          = "stool_aceticAcid",
                          "Propionic acid (Stool)"       = "stool_propionicAcid",
                          "Butyric acid (Stool)"         = "stool_butyricAcid",
                          "Valeric acid (Stool)"         = "stool_valericAcid",
                          "Hexanoic acid (Stool)"        = "stool_hexanoicAcid")
BCFA_markers         <- c("Iso-butyric acid (Stool)"     = "stool_isobutyricAcid",
                          "Iso-valeric acid (Stool)"     = "stool_isovalericAcid",
                          "2-Methylbutyric acid (Stool)" = "stool_methylbutyricAcid2")

col_vector <- c("scfa" = "#3889E5",
                "bcfa" = "#9769C9")
```


```{r helper functions}
# descriptive plot
plot_variable <- function(dat, variable, cut_extremeValues = TRUE) {
  colnames(dat)[colnames(dat) == variable] <- "y"
  
  ylim <- c(min(dat$y, na.rm = TRUE), quantile(dat$y, probs = ifelse(cut_extremeValues, 0.95, 1), na.rm = TRUE))
  
  gg_time <- dat %>% 
    ggplot(aes(x = timepoint, y = y, fill = group)) +
    geom_boxplot() +
    facet_wrap(~ group) +
    ggtitle(paste(variable, "over time"), subtitle = ifelse(cut_extremeValues, "upper 5% of observations cut off", "")) +
    ylab(variable) + ylim(ylim) +
    theme(legend.position = "none",
          axis.title.x    = element_blank())
  
  gg_t1vst3 <- dat %>% 
    filter(timepoint %in% c("T1", "T3")) %>% 
    ggplot(aes(x = timepoint, y = y)) +
    geom_boxplot(fill = "gray95") +
    ggtitle("T1 vs. T3", subtitle = " ") +
    ylim(ylim) +
    theme(axis.title.y = element_blank(),
          axis.text.y  = element_blank())
  
  gg_time + gg_t1vst3 + patchwork::plot_layout(widths = c(.8, .2))
}

# estimate a regression model with mgcv
estimate_model <- function(dat, y, family = gaussian(), transformation = identity, interaction_string = "", estimation_method = "REML") {
  
  dat[[y]] <- dat[[y]] %>% transformation()
  
  covar_string <- paste0("~ intervention +",
                         ifelse(interaction_string == "", "", paste(interaction_string, "+")),
                         "phase + gender + age_centered50 + bmi_centered25 + s(participant_id, bs = 're')")
  
  paste(y, covar_string) %>% formula() %>% 
    gam(method = estimation_method, data = dat, family = family)
}


# create a standard residual plot
plot_residuals <- function(gam_model, title = "") {
  
  plot_dat <- data.frame(predictions = gam_model$fitted.values,
                         residuals   = gam_model$residuals)
  
  ggplot(plot_dat, aes(x = predictions, y = residuals)) +
    geom_hline(yintercept = 0, col = "firebrick", lty = 2) +
    geom_point() +
    ylim(c(-1,1) * max(abs(plot_dat$residuals))) +
    ggtitle(title)
}
```

```{r helper function model comparison}
# estimate both a normal distribution and a Gamma distribution model and return
# a list of comparison plots and tables
compare_models <- function(y) {
  
  # also reestimate the interaction models once each with changed reference category,
  # to eventually easily retrieve the individual intervention effects
  dat_model_changedRef <- dat_model %>% 
    mutate(gender  = relevel(gender,  ref = "W"),
           age_cat = relevel(age_cat, ref = "50-69"),
           bmi_cat = relevel(bmi_cat, ref = "[25, 31]"))
  
  # all returned final models are estimated using REML,
  # all models used for testing potential interaction effects via likelihood ratio tests are estimated using ML
  
  m_gaussian           <- estimate_model(dat_model, y, family = gaussian())
  m_gaussian_ML        <- estimate_model(dat_model, y, family = gaussian(), estimation_method = "ML")
  m_gaussian_sexInt    <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:gender")
  m_gaussian_sexInt_ML <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:gender", estimation_method = "ML")
  m_gaussian_ageInt    <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:age_cat")
  m_gaussian_ageInt_ML <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:age_cat", estimation_method = "ML")
  m_gaussian_bmiInt    <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:bmi_cat")
  m_gaussian_bmiInt_ML <- estimate_model(dat_model, y, family = gaussian(),                       interaction_string = "intervention:bmi_cat", estimation_method = "ML")
  m_gaussian_sexInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(),    interaction_string = "intervention:gender")
  m_gaussian_ageInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(),    interaction_string = "intervention:age_cat")
  m_gaussian_bmiInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(),    interaction_string = "intervention:bmi_cat")
  
  m_gaussianLog           <- estimate_model(dat_model, y, family = gaussian(), transformation = log)
  m_gaussianLog_ML        <- estimate_model(dat_model, y, family = gaussian(), transformation = log, estimation_method = "ML")
  m_gaussianLog_sexInt    <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:gender")
  m_gaussianLog_sexInt_ML <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:gender", estimation_method = "ML")
  m_gaussianLog_ageInt    <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:age_cat")
  m_gaussianLog_ageInt_ML <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:age_cat", estimation_method = "ML")
  m_gaussianLog_bmiInt    <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:bmi_cat")
  m_gaussianLog_bmiInt_ML <- estimate_model(dat_model, y, family = gaussian(), transformation = log, interaction_string = "intervention:bmi_cat", estimation_method = "ML")
  m_gaussianLog_sexInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(), transformation = log, interaction_string = "intervention:gender")
  m_gaussianLog_ageInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(), transformation = log, interaction_string = "intervention:age_cat")
  m_gaussianLog_bmiInt_changedRef <- estimate_model(dat_model_changedRef, y, family = gaussian(), transformation = log, interaction_string = "intervention:bmi_cat")

  m_gamma              <- estimate_model(dat_model, y, family = Gamma(link = "log"))
  m_gamma_ML           <- estimate_model(dat_model, y, family = Gamma(link = "log"), estimation_method = "ML")
  m_gamma_sexInt       <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:gender")
  m_gamma_sexInt_ML    <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:gender", estimation_method = "ML")
  m_gamma_ageInt       <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:age_cat")
  m_gamma_ageInt_ML    <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:age_cat", estimation_method = "ML")
  m_gamma_bmiInt       <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:bmi_cat")
  m_gamma_bmiInt_ML    <- estimate_model(dat_model, y, family = Gamma(link = "log"),              interaction_string = "intervention:bmi_cat", estimation_method = "ML")
  m_gamma_sexInt_changedRef <- estimate_model(dat_model_changedRef, y, family = Gamma(link = "log"), interaction_string = "intervention:gender")
  m_gamma_ageInt_changedRef <- estimate_model(dat_model_changedRef, y, family = Gamma(link = "log"), interaction_string = "intervention:age_cat")
  m_gamma_bmiInt_changedRef <- estimate_model(dat_model_changedRef, y, family = Gamma(link = "log"), interaction_string = "intervention:bmi_cat")
  
  # 1) table of explained deviances and tests against the base model
  p_gs_sexInt    <- anova.gam(m_gaussian_ML,    m_gaussian_sexInt_ML,    test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gs_ageInt    <- anova.gam(m_gaussian_ML,    m_gaussian_ageInt_ML,    test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gs_bmiInt    <- anova.gam(m_gaussian_ML,    m_gaussian_bmiInt_ML,    test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gsLog_sexInt <- anova.gam(m_gaussianLog_ML, m_gaussianLog_sexInt_ML, test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gsLog_ageInt <- anova.gam(m_gaussianLog_ML, m_gaussianLog_ageInt_ML, test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gsLog_bmiInt <- anova.gam(m_gaussianLog_ML, m_gaussianLog_bmiInt_ML, test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gm_sexInt    <- anova.gam(m_gamma_ML,       m_gamma_sexInt_ML,       test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gm_ageInt    <- anova.gam(m_gamma_ML,       m_gamma_ageInt_ML,       test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  p_gm_bmiInt    <- anova.gam(m_gamma_ML,       m_gamma_bmiInt_ML,       test = "Chisq")$`Pr(>Chi)`[2] %>% round(4)
  
  # in case the simpler model has lower (= better) deviance, no p-values is calculated by 'anova.gam'
  p_gs_sexInt    <- ifelse(is.na(p_gs_sexInt),    0.9999, p_gs_sexInt)
  p_gs_ageInt    <- ifelse(is.na(p_gs_ageInt),    0.9999, p_gs_ageInt)
  p_gs_bmiInt    <- ifelse(is.na(p_gs_bmiInt),    0.9999, p_gs_bmiInt)
  p_gsLog_sexInt <- ifelse(is.na(p_gsLog_sexInt), 0.9999, p_gsLog_sexInt)
  p_gsLog_ageInt <- ifelse(is.na(p_gsLog_ageInt), 0.9999, p_gsLog_ageInt)
  p_gsLog_bmiInt <- ifelse(is.na(p_gsLog_bmiInt), 0.9999, p_gsLog_bmiInt)
  p_gm_sexInt    <- ifelse(is.na(p_gm_sexInt),    0.9999, p_gm_sexInt)
  p_gm_ageInt    <- ifelse(is.na(p_gm_ageInt),    0.9999, p_gm_ageInt)
  p_gm_bmiInt    <- ifelse(is.na(p_gm_bmiInt),    0.9999, p_gm_bmiInt)
  
  tab_dev <- data.frame("Model"             = c("Raw intervention effect", "Intervention effect by gender", "Intervention effect by age group", "Intervention effect by BMI group"),
                        "Gaussian"          = c(paste0(round(100 * summary(m_gaussian_ML)$dev.expl,           1), "%"),
                                                paste0(round(100 * summary(m_gaussian_sexInt_ML)$dev.expl,    1), "% (p = ", p_gs_sexInt, ")"),
                                                paste0(round(100 * summary(m_gaussian_ageInt_ML)$dev.expl,    1), "% (p = ", p_gs_ageInt, ")"),
                                                paste0(round(100 * summary(m_gaussian_bmiInt_ML)$dev.expl,    1), "% (p = ", p_gs_bmiInt, ")")),
                        "Gaussian_logTrafo" = c(paste0(round(100 * summary(m_gaussianLog_ML)$dev.expl,        1), "%"),
                                                paste0(round(100 * summary(m_gaussianLog_sexInt_ML)$dev.expl, 1), "% (p = ", p_gsLog_sexInt, ")"),
                                                paste0(round(100 * summary(m_gaussianLog_ageInt_ML)$dev.expl, 1), "% (p = ", p_gsLog_ageInt, ")"),
                                                paste0(round(100 * summary(m_gaussianLog_bmiInt_ML)$dev.expl, 1), "% (p = ", p_gsLog_bmiInt, ")")),
                        "Gamma"             = c(paste0(round(100 * summary(m_gamma_ML)$dev.expl,              1), "%"),
                                                paste0(round(100 * summary(m_gamma_sexInt_ML)$dev.expl,       1), "% (p = ", p_gm_sexInt, ")"),
                                                paste0(round(100 * summary(m_gamma_ageInt_ML)$dev.expl,       1), "% (p = ", p_gm_ageInt, ")"),
                                                paste0(round(100 * summary(m_gamma_bmiInt_ML)$dev.expl,       1), "% (p = ", p_gm_bmiInt, ")")))
  colnames(tab_dev)[2:4] <- c("Gaussian", "Gaussian (log trafo)", "Gamma")
  
  # 2) table of intervention effects (to judge the potential interactions)
  est_gs <- summary(m_gaussian)$p.table[,"Estimate"]
  est_gm <- summary(m_gamma)$p.table[,"Estimate"]
  tab_est_noInt <- data.frame("Model"    = "Raw intervention effect",
                              "Effect"   = c("Fresh vs. baseline", "Pasteurized vs. baseline"),
                              "Gaussian" = est_gs[c("interventionFresh", "interventionPasteurized")] %>% unname() %>% round(1),
                              "Gamma"    = est_gm[c("interventionFresh", "interventionPasteurized")] %>% unname() %>% round(3))

  est_gsA <- summary(m_gaussian_ageInt)$p.table[,"Estimate"]
  est_gmA <- summary(m_gamma_ageInt)$p.table[,"Estimate"]
  tab_est_ageInt <- data.frame("Model"    = "Intervention effect by age group",
                               "Effect"   = paste(rep(c("Fresh vs. baseline", "Pasteurized vs. baseline"), each  = 2),
                                                  rep(c("- age 21-49", "- age 50-69"),                     times = 2)),
                               "Gaussian" = c(est_gsA[["interventionFresh"]],
                                              est_gsA[["interventionFresh"]] + est_gsA[["interventionFresh:age_cat50-69"]] - est_gsA[["interventionBaseline:age_cat50-69"]],
                                              est_gsA[["interventionPasteurized"]],
                                              est_gsA[["interventionPasteurized"]] + est_gsA[["interventionPasteurized:age_cat50-69"]] - est_gsA[["interventionBaseline:age_cat50-69"]]),
                               "Gamma"    = c(est_gmA[["interventionFresh"]],
                                              est_gmA[["interventionFresh"]] + est_gmA[["interventionFresh:age_cat50-69"]] - est_gmA[["interventionBaseline:age_cat50-69"]],
                                              est_gmA[["interventionPasteurized"]],
                                              est_gmA[["interventionPasteurized"]] + est_gmA[["interventionPasteurized:age_cat50-69"]] - est_gmA[["interventionBaseline:age_cat50-69"]]))

  est_gsB <- summary(m_gaussian_bmiInt)$p.table[,"Estimate"]
  est_gmB <- summary(m_gamma_bmiInt)$p.table[,"Estimate"]
  tab_est_bmiInt <- data.frame("Model" = "Intervention effect by BMI group",
                               "Effect"   = paste(rep(c("Fresh vs. baseline", "Pasteurized vs. baseline"), each  = 2),
                                                  rep(c("- BMI [18, 25)", "- BMI [25, 31]",                times = 2))),
                               "Gaussian" = c(est_gsB[["interventionFresh"]],
                                              est_gsB[["interventionFresh"]] + est_gsB[["interventionFresh:bmi_cat[25, 31]"]] - est_gsB[["interventionBaseline:bmi_cat[25, 31]"]],
                                              est_gsB[["interventionPasteurized"]],
                                              est_gsB[["interventionPasteurized"]] + est_gsB[["interventionPasteurized:bmi_cat[25, 31]"]] - est_gsB[["interventionBaseline:bmi_cat[25, 31]"]]),
                               "Gamma"    = c(est_gmB[["interventionFresh"]],
                                              est_gmB[["interventionFresh"]] + est_gmB[["interventionFresh:bmi_cat[25, 31]"]] - est_gmB[["interventionBaseline:bmi_cat[25, 31]"]],
                                              est_gmB[["interventionPasteurized"]],
                                              est_gmB[["interventionPasteurized"]] + est_gmB[["interventionPasteurized:bmi_cat[25, 31]"]] - est_gmB[["interventionBaseline:bmi_cat[25, 31]"]]))

  tab_effects <- tab_est_noInt %>% dplyr::bind_rows(tab_est_ageInt) %>% dplyr::bind_rows(tab_est_bmiInt) %>%
    mutate(Gaussian = round(Gaussian, 2),
           Gamma    = round(Gamma, 3))
  
  # 3) residual plots
  gg_residuals <- plot_residuals(m_gaussian, "Gaussian\nwithout interaction") +
    plot_residuals(m_gaussianLog, "Gaussian with log trafo\nwithout interaction") +
    plot_residuals(m_gamma, "Gamma\nwithout interaction") +
    patchwork::plot_layout(nrow = 1)
  
  # 4) QQ plots
  qq_gaussian    <- m_gaussian %>% mgcViz::getViz() %>% mgcViz::qq.gamViz() +
    ggtitle("Gaussian") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
  qq_gaussianLog <- m_gaussianLog %>% mgcViz::getViz() %>% mgcViz::qq.gamViz() +
    ggtitle("Gaussian (log trafo)") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
  qq_gamma       <- m_gamma    %>% mgcViz::getViz() %>% mgcViz::qq.gamViz() +
    ggtitle("Gamma") + theme_minimal() + theme(plot.title = element_text(hjust = 0.5))
  
  
  list("model_gaussian"           = m_gaussian,
       "model_gaussian_sexInt"    = m_gaussian_sexInt,
       "model_gaussian_ageInt"    = m_gaussian_ageInt,
       "model_gaussian_bmiInt"    = m_gaussian_bmiInt,
       "model_gaussian_sexInt_changedRef" = m_gaussian_sexInt_changedRef,
       "model_gaussian_ageInt_changedRef" = m_gaussian_ageInt_changedRef,
       "model_gaussian_bmiInt_changedRef" = m_gaussian_bmiInt_changedRef,
       "model_gaussianLog"        = m_gaussianLog,
       "model_gaussianLog_sexInt" = m_gaussianLog_sexInt,
       "model_gaussianLog_ageInt" = m_gaussianLog_ageInt,
       "model_gaussianLog_bmiInt" = m_gaussianLog_bmiInt,
       "model_gaussianLog_sexInt_changedRef" = m_gaussianLog_sexInt_changedRef,
       "model_gaussianLog_ageInt_changedRef" = m_gaussianLog_ageInt_changedRef,
       "model_gaussianLog_bmiInt_changedRef" = m_gaussianLog_bmiInt_changedRef,
       "model_gamma"              = m_gamma,
       "model_gamma_sexInt"       = m_gamma_sexInt,
       "model_gamma_ageInt"       = m_gamma_ageInt,
       "model_gamma_bmiInt"       = m_gamma_bmiInt,
       "model_gamma_sexInt_changedRef" = m_gamma_sexInt_changedRef,
       "model_gamma_ageInt_changedRef" = m_gamma_ageInt_changedRef,
       "model_gamma_bmiInt_changedRef" = m_gamma_bmiInt_changedRef,
       "tab_explainedDeviances"   = tab_dev,
       "tab_effects"              = tab_effects,
       "gg_residuals"             = gg_residuals,
       "qq_gaussian"              = qq_gaussian,
       "qq_gaussianLog"           = qq_gaussianLog,
       "qq_gamma"                 = qq_gamma) %>% 
    return()
}
```


# SCRF - 2-Methylbutyric acid

## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_methylbutyricAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_methylbutyric_acid2 <- compare_models("stool_methylbutyricAcid")

outputList_methylbutyric_acid2$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_methylbutyric_acid2$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_methylbutyric_acid2$gg_residuals
mgcViz::gridPrint(outputList_methylbutyric_acid2$qq_gaussian, outputList_methylbutyric_acid2$qq_gaussianLog, outputList_methylbutyric_acid2$qq_gamma, nrow = 1)
```

## Final model (log Gaussian)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_methylbutyric_acid2        <- outputList_methylbutyric_acid2$model_gaussianLog

APCtools:::extract_summary_linearEffects(m_methylbutyric_acid2) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_methylbutyric_acid2)

# save the interaction models for later use for visualization
m_methylbutyric_acid2_sexInt            <- outputList_methylbutyric_acid2$model_gaussianLog_sexInt
m_methylbutyric_acid2_sexInt_changedRef <- outputList_methylbutyric_acid2$model_gaussianLog_sexInt_changedRef
m_methylbutyric_acid2_ageInt            <- outputList_methylbutyric_acid2$model_gaussianLog_ageInt
m_methylbutyric_acid2_ageInt_changedRef <- outputList_methylbutyric_acid2$model_gaussianLog_ageInt_changedRef
m_methylbutyric_acid2_bmiInt            <- outputList_methylbutyric_acid2$model_gaussianLog_bmiInt
m_methylbutyric_acid2_bmiInt_changedRef <- outputList_methylbutyric_acid2$model_gaussianLog_bmiInt_changedRef
```


# SCRF - Acetic acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_aceticAcid) + 3*sd(dat$stool_aceticAcid)
# 
# dat <- dat %>% 
#   mutate(stool_aceticAcid_trimmed = case_when(stool_aceticAcid < upper_threshold ~ stool_aceticAcid,
#                                               TRUE                               ~ upper_threshold))
# dat_model <- dat_model %>% 
#   mutate(stool_aceticAcid_trimmed = case_when(stool_aceticAcid < upper_threshold ~ stool_aceticAcid,
#                                               TRUE                               ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_aceticAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_acetic_acid <- compare_models("stool_aceticAcid")

outputList_acetic_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_acetic_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_acetic_acid$gg_residuals
mgcViz::gridPrint(outputList_acetic_acid$qq_gaussian, outputList_acetic_acid$qq_gaussianLog, outputList_acetic_acid$qq_gamma, nrow = 1)
```

## Final model (Gamma)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_acetic_acid        <- outputList_acetic_acid$model_gamma

APCtools:::extract_summary_linearEffects(m_acetic_acid) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_acetic_acid)

# save the interaction models for later use for visualization
m_acetic_acid_sexInt            <- outputList_acetic_acid$model_gamma_sexInt
m_acetic_acid_sexInt_changedRef <- outputList_acetic_acid$model_gamma_sexInt_changedRef
m_acetic_acid_ageInt            <- outputList_acetic_acid$model_gamma_ageInt
m_acetic_acid_ageInt_changedRef <- outputList_acetic_acid$model_gamma_ageInt_changedRef
m_acetic_acid_bmiInt            <- outputList_acetic_acid$model_gamma_bmiInt
m_acetic_acid_bmiInt_changedRef <- outputList_acetic_acid$model_gamma_bmiInt_changedRef
```


# SCRF - Butyric acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_butyricAcid) + 3*sd(dat$stool_butyricAcid)
# 
# dat <- dat %>%
#   mutate(stool_butyricAcid_trimmed = case_when(stool_butyricAcid < upper_threshold ~ stool_butyricAcid,
#                                                TRUE                               ~ upper_threshold))
# dat_model <- dat_model %>%
#   mutate(stool_butyricAcid_trimmed = case_when(stool_butyricAcid < upper_threshold ~ stool_butyricAcid,
#                                                TRUE                               ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_butyricAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_butyric_acid <- compare_models("stool_butyricAcid")

outputList_butyric_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_butyric_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_butyric_acid$gg_residuals
mgcViz::gridPrint(outputList_butyric_acid$qq_gaussian, outputList_butyric_acid$qq_gaussianLog, outputList_butyric_acid$qq_gamma, nrow = 1)
```

## Final model (Gamma)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_butyric_acid        <- outputList_butyric_acid$model_gamma

APCtools:::extract_summary_linearEffects(m_butyric_acid) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_butyric_acid)

# save the interaction models for later use for visualization
m_butyric_acid_sexInt            <- outputList_butyric_acid$model_gamma_sexInt
m_butyric_acid_sexInt_changedRef <- outputList_butyric_acid$model_gamma_sexInt_changedRef
m_butyric_acid_ageInt            <- outputList_butyric_acid$model_gamma_ageInt
m_butyric_acid_ageInt_changedRef <- outputList_butyric_acid$model_gamma_ageInt_changedRef
m_butyric_acid_bmiInt            <- outputList_butyric_acid$model_gamma_bmiInt
m_butyric_acid_bmiInt_changedRef <- outputList_butyric_acid$model_gamma_bmiInt_changedRef
```


# SCRF - Hexanoic acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_hexanoicAcid) + 3*sd(dat$stool_hexanoicAcid)
# 
# dat <- dat %>%
#   mutate(stool_hexanoicAcid_trimmed = case_when(stool_hexanoicAcid < upper_threshold ~ stool_hexanoicAcid,
#                                                 TRUE                               ~ upper_threshold))
# dat_model <- dat_model %>%
#   mutate(stool_hexanoicAcid_trimmed = case_when(stool_hexanoicAcid < upper_threshold ~ stool_hexanoicAcid,
#                                                 TRUE                               ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_hexanoicAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_hexanoic_acid <- compare_models("stool_hexanoicAcid")

outputList_hexanoic_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_hexanoic_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_hexanoic_acid$gg_residuals
mgcViz::gridPrint(outputList_hexanoic_acid$qq_gaussian, outputList_hexanoic_acid$qq_gaussianLog, outputList_hexanoic_acid$qq_gamma, nrow = 1)
```

## Final model (Gaussian)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_hexanoic_acid        <- outputList_hexanoic_acid$model_gaussian

APCtools:::extract_summary_linearEffects(m_hexanoic_acid) %>% 
  mutate(pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_hexanoic_acid)

# save the interaction models for later use for visualization
m_hexanoic_acid_sexInt            <- outputList_hexanoic_acid$model_gaussian_sexInt
m_hexanoic_acid_sexInt_changedRef <- outputList_hexanoic_acid$model_gaussian_sexInt_changedRef
m_hexanoic_acid_ageInt            <- outputList_hexanoic_acid$model_gaussian_ageInt
m_hexanoic_acid_ageInt_changedRef <- outputList_hexanoic_acid$model_gaussian_ageInt_changedRef
m_hexanoic_acid_bmiInt            <- outputList_hexanoic_acid$model_gaussian_bmiInt
m_hexanoic_acid_bmiInt_changedRef <- outputList_hexanoic_acid$model_gaussian_bmiInt_changedRef
```


# SCRF - Isobutyric acid

## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_isobutyricAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_isobutyric_acid <- compare_models("stool_isobutyricAcid")

outputList_isobutyric_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_isobutyric_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_isobutyric_acid$gg_residuals
mgcViz::gridPrint(outputList_isobutyric_acid$qq_gaussian, outputList_isobutyric_acid$qq_gaussianLog, outputList_isobutyric_acid$qq_gamma, nrow = 1)
```

## Final model (Gamma)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_isobutyric_acid        <- outputList_isobutyric_acid$model_gamma

APCtools:::extract_summary_linearEffects(m_isobutyric_acid) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_isobutyric_acid)

# save the interaction models for later use for visualization
m_isobutyric_acid_sexInt            <- outputList_isobutyric_acid$model_gamma_sexInt
m_isobutyric_acid_sexInt_changedRef <- outputList_isobutyric_acid$model_gamma_sexInt_changedRef
m_isobutyric_acid_ageInt            <- outputList_isobutyric_acid$model_gamma_ageInt
m_isobutyric_acid_ageInt_changedRef <- outputList_isobutyric_acid$model_gamma_ageInt_changedRef
m_isobutyric_acid_bmiInt            <- outputList_isobutyric_acid$model_gamma_bmiInt
m_isobutyric_acid_bmiInt_changedRef <- outputList_isobutyric_acid$model_gamma_bmiInt_changedRef
```


# SCRF - Isovaleric acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_isovalericAcid) + 3*sd(dat$stool_isovalericAcid)
# 
# dat <- dat %>%
#   mutate(stool_isovalericAcid_trimmed = case_when(stool_isovalericAcid < upper_threshold ~ stool_isovalericAcid,
#                                                   TRUE                               ~ upper_threshold))
# dat_model <- dat_model %>%
#   mutate(stool_isovalericAcid_trimmed = case_when(stool_isovalericAcid < upper_threshold ~ stool_isovalericAcid,
#                                                   TRUE                               ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_isovalericAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_isovaleric_acid <- compare_models("stool_isovalericAcid")

outputList_isovaleric_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_isovaleric_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_isovaleric_acid$gg_residuals
mgcViz::gridPrint(outputList_isovaleric_acid$qq_gaussian, outputList_isovaleric_acid$qq_gaussianLog, outputList_isovaleric_acid$qq_gamma, nrow = 1)
```

## Final model (Gamma)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_isovaleric_acid        <- outputList_isovaleric_acid$model_gamma

APCtools:::extract_summary_linearEffects(m_isovaleric_acid) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_isovaleric_acid)

# save the interaction models for later use for visualization
m_isovaleric_acid_sexInt            <- outputList_isovaleric_acid$model_gamma_sexInt
m_isovaleric_acid_sexInt_changedRef <- outputList_isovaleric_acid$model_gamma_sexInt_changedRef
m_isovaleric_acid_ageInt            <- outputList_isovaleric_acid$model_gamma_ageInt
m_isovaleric_acid_ageInt_changedRef <- outputList_isovaleric_acid$model_gamma_ageInt_changedRef
m_isovaleric_acid_bmiInt            <- outputList_isovaleric_acid$model_gamma_bmiInt
m_isovaleric_acid_bmiInt_changedRef <- outputList_isovaleric_acid$model_gamma_bmiInt_changedRef
```


# SCRF - Propionic acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_propionicAcid) + 3*sd(dat$stool_propionicAcid)
# 
# dat <- dat %>%
#   mutate(stool_propionicAcid_trimmed = case_when(stool_propionicAcid < upper_threshold ~ stool_propionicAcid,
#                                                  TRUE                               ~ upper_threshold))
# dat_model <- dat_model %>%
#   mutate(stool_propionicAcid_trimmed = case_when(stool_propionicAcid < upper_threshold ~ stool_propionicAcid,
#                                                  TRUE                               ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_propionicAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_propionic_acid <- compare_models("stool_propionicAcid")

outputList_propionic_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_propionic_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_propionic_acid$gg_residuals
mgcViz::gridPrint(outputList_propionic_acid$qq_gaussian, outputList_propionic_acid$qq_gaussianLog, outputList_propionic_acid$qq_gamma, nrow = 1)
```

## Final model (Gamma)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_propionic_acid        <- outputList_propionic_acid$model_gamma

APCtools:::extract_summary_linearEffects(m_propionic_acid) %>% 
  mutate(coef_exp = exp(coef),
         se_exp   = exp(se),
         CI_lower_exp = exp(CI_lower),
         CI_upper_exp = exp(CI_upper),
         pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_propionic_acid)

# save the interaction models for later use for visualization
m_propionic_acid_sexInt            <- outputList_propionic_acid$model_gamma_sexInt
m_propionic_acid_sexInt_changedRef <- outputList_propionic_acid$model_gamma_sexInt_changedRef
m_propionic_acid_ageInt            <- outputList_propionic_acid$model_gamma_ageInt
m_propionic_acid_ageInt_changedRef <- outputList_propionic_acid$model_gamma_ageInt_changedRef
m_propionic_acid_bmiInt            <- outputList_propionic_acid$model_gamma_bmiInt
m_propionic_acid_bmiInt_changedRef <- outputList_propionic_acid$model_gamma_bmiInt_changedRef
```


# SCRF - Valeric acid

## Sensitivity analysis results

All below results were estimated on the observed data, without specific outlier
exclusion / handling.
To check the influence of outliers on the results, the below models were re-estimated
on the dataset in which outliers (= points above the threshold 'mean + 3*sd') were
trimmed to the treshold 'mean + 3*sd'.
These sensitivity results are not comprised in this document.
The results based on the trimmed data were not substantially different from the original results.

```{r}
# # trim outliers to mean + 3*sd
# upper_threshold <- median(dat$stool_valericAcid) + 3*sd(dat$stool_valericAcid)
# 
# dat <- dat %>%
#   mutate(stool_valericAcid_trimmed = case_when(stool_valericAcid < upper_threshold ~ stool_valericAcid,
#                                                TRUE                                ~ upper_threshold))
# dat_model <- dat_model %>%
#   mutate(stool_valericAcid_trimmed = case_when(stool_valericAcid < upper_threshold ~ stool_valericAcid,
#                                                TRUE                                ~ upper_threshold))
```


## Description

```{r}
#| fig.width: 8
#| fig.height: 5
#| warning: false

plot_variable(dat, "stool_valericAcid", cut_extremeValues = FALSE)
```

## Model comparison

Share of deviance explained (including p-values for testing the respective interaction model against the model without interaction):

```{r}
#| fig.width: 8
#| fig.height: 4

outputList_valeric_acid <- compare_models("stool_valericAcid")

outputList_valeric_acid$tab_explainedDeviances %>% kable() %>% kable_styling()
# outputList_valeric_acid$tab_effects %>% kable() %>% kable_styling() %>%
#   group_rows("Intervention by age", 3, 6) %>% group_rows("Intervention by BMI", 7, 10)
outputList_valeric_acid$gg_residuals
mgcViz::gridPrint(outputList_valeric_acid$qq_gaussian, outputList_valeric_acid$qq_gaussianLog, outputList_valeric_acid$qq_gamma, nrow = 1)
```

## Final model (Gaussian)

##### Estimated effects

```{r}
#| fig.width: 8
#| fig.height: 4

m_valeric_acid        <- outputList_valeric_acid$model_gaussian

APCtools:::extract_summary_linearEffects(m_valeric_acid) %>% 
  mutate(pvalue       = case_when(param == "(Intercept)" ~ "-",
                                  TRUE                   ~ pvalue %>% round(4) %>% as.character()),
         across(!contains("param") & !contains("pvalue"), function(x) { round(x, 3) })) %>% 
  select(!contains("pvalue"), pvalue) %>% 
  kable() %>% 
  kable_styling()
APCtools::plot_linearEffects(m_valeric_acid)

# save the interaction models for later use for visualization
m_valeric_acid_sexInt            <- outputList_valeric_acid$model_gaussian_sexInt
m_valeric_acid_sexInt_changedRef <- outputList_valeric_acid$model_gaussian_sexInt_changedRef
m_valeric_acid_ageInt            <- outputList_valeric_acid$model_gaussian_ageInt
m_valeric_acid_ageInt_changedRef <- outputList_valeric_acid$model_gaussian_ageInt_changedRef
m_valeric_acid_bmiInt            <- outputList_valeric_acid$model_gaussian_bmiInt
m_valeric_acid_bmiInt_changedRef <- outputList_valeric_acid$model_gaussian_bmiInt_changedRef
```


# Overview table of all models

```{r}
# create a joint vector of all variables
model_names <- c(SCFA_markers, BCFA_markers)
# create a model list
model_list <- list("Acetic acid (Stool)"          = m_acetic_acid,
                   "Propionic acid (Stool)"       = m_propionic_acid,
                   "Butyric acid (Stool)"         = m_butyric_acid,
                   "Valeric acid (Stool)"         = m_valeric_acid,
                   "Hexanoic acid (Stool)"        = m_hexanoic_acid,
                   "Iso-butyric acid (Stool)"     = m_isobutyric_acid,
                   "Iso-valeric acid (Stool)"     = m_isovaleric_acid,
                   "2-Methylbutyric acid (Stool)" = m_methylbutyric_acid2)
model_sexInt_list <- list("Acetic acid (Stool)"          = list(m_acetic_acid_sexInt, m_acetic_acid_sexInt_changedRef),
                          "Propionic acid (Stool)"       = list(m_propionic_acid_sexInt, m_propionic_acid_sexInt_changedRef),
                          "Butyric acid (Stool)"         = list(m_butyric_acid_sexInt, m_butyric_acid_sexInt_changedRef),
                          "Valeric acid (Stool)"         = list(m_valeric_acid_sexInt, m_valeric_acid_sexInt_changedRef),
                          "Hexanoic acid (Stool)"        = list(m_hexanoic_acid_sexInt, m_hexanoic_acid_sexInt_changedRef),
                          "Iso-butyric acid (Stool)"     = list(m_isobutyric_acid_sexInt, m_isobutyric_acid_sexInt_changedRef),
                          "Iso-valeric acid (Stool)"     = list(m_isovaleric_acid_sexInt, m_isovaleric_acid_sexInt_changedRef),
                          "2-Methylbutyric acid (Stool)" = list(m_methylbutyric_acid2_sexInt, m_methylbutyric_acid2_sexInt_changedRef))
model_ageInt_list <- list("Acetic acid (Stool)"          = list(m_acetic_acid_ageInt, m_acetic_acid_ageInt_changedRef),
                          "Propionic acid (Stool)"       = list(m_propionic_acid_ageInt, m_propionic_acid_ageInt_changedRef),
                          "Butyric acid (Stool)"         = list(m_butyric_acid_ageInt, m_butyric_acid_ageInt_changedRef),
                          "Valeric acid (Stool)"         = list(m_valeric_acid_ageInt, m_valeric_acid_ageInt_changedRef),
                          "Hexanoic acid (Stool)"        = list(m_hexanoic_acid_ageInt, m_hexanoic_acid_ageInt_changedRef),
                          "Iso-butyric acid (Stool)"     = list(m_isobutyric_acid_ageInt, m_isobutyric_acid_ageInt_changedRef),
                          "Iso-valeric acid (Stool)"     = list(m_isovaleric_acid_ageInt, m_isovaleric_acid_ageInt_changedRef),
                          "2-Methylbutyric acid (Stool)" = list(m_methylbutyric_acid2_ageInt, m_methylbutyric_acid2_ageInt_changedRef))
model_bmiInt_list <- list("Acetic acid (Stool)"          = list(m_acetic_acid_bmiInt, m_acetic_acid_bmiInt_changedRef),
                          "Propionic acid (Stool)"       = list(m_propionic_acid_bmiInt, m_propionic_acid_bmiInt_changedRef),
                          "Butyric acid (Stool)"         = list(m_butyric_acid_bmiInt, m_butyric_acid_bmiInt_changedRef),
                          "Valeric acid (Stool)"         = list(m_valeric_acid_bmiInt, m_valeric_acid_bmiInt_changedRef),
                          "Hexanoic acid (Stool)"        = list(m_hexanoic_acid_bmiInt, m_hexanoic_acid_bmiInt_changedRef),
                          "Iso-butyric acid (Stool)"     = list(m_isobutyric_acid_bmiInt, m_isobutyric_acid_bmiInt_changedRef),
                          "Iso-valeric acid (Stool)"     = list(m_isovaleric_acid_bmiInt, m_isovaleric_acid_bmiInt_changedRef),
                          "2-Methylbutyric acid (Stool)" = list(m_methylbutyric_acid2_bmiInt, m_methylbutyric_acid2_bmiInt_changedRef))
# create a lookup list for the final model type
modelType_list <- list("Acetic acid (Stool)"          = "Gamma",
                       "Propionic acid (Stool)"       = "Gamma",
                       "Butyric acid (Stool)"         = "Gamma",
                       "Valeric acid (Stool)"         = "Gaussian",
                       "Hexanoic acid (Stool)"        = "Gaussian",
                       "Iso-butyric acid (Stool)"     = "Gamma",
                       "Iso-valeric acid (Stool)"     = "Gamma",
                       "2-Methylbutyric acid (Stool)" = "log Gaussian")


# overview table
data.frame(Model              = names(model_list),
           Model_type         = unlist(modelType_list, use.names = FALSE),
           Deviance_explained = sapply(model_list, function(x) { paste0(round(100 * summary(x)$dev.expl, 1), "%") }, USE.NAMES = FALSE) %>% unname()) %>% 
  kable() %>% 
  kable_styling()
```


# Joint plot of all effects

```{r prepare all model results in one dataset}
# create a results table with all effects
modelResults_list <- lapply(names(model_names), function(x) {
  
  variable   <- model_names[x] %>% unname()
  model      <- model_list[[x]]
  model_type <- modelType_list[[x]]
  
  log_wasApplied <- (model_type %in% c("Gamma", "log Gaussian"))
  
  results_dat <- summary(model)$p.table %>% 
    as.data.frame() %>% 
    mutate(parameter   = row.names(.),
           param_group = case_when(parameter %in% c("interventionFresh", "interventionPasteurized") ~ "intervention",
                                   grepl("carry_over", parameter)                                   ~ "carry-over",
                                   grepl("phase", parameter)                                        ~ "phase",
                                   grepl("gender", parameter)                                       ~ "gender",
                                   grepl("bmi", parameter)                                          ~ "BMI",
                                   grepl("age", parameter)                                          ~ "age",
                                   TRUE                                                             ~ parameter)) %>% 
    dplyr::rename(estimate   = "Estimate",
                  se         = "Std. Error",
                  pvalue     = "Pr(>|t|)") %>% 
    mutate(model      = x,
           model_type = model_type,
           CI_lower   = estimate - qnorm(0.975) * se,
           CI_upper   = estimate + qnorm(0.975) * se)
  
  if (!log_wasApplied) { # standardize the effects by the standard deviation of the response variable
    results_dat <- results_dat %>% 
      mutate(y_sd = sd(dat_model[[variable]], na.rm = TRUE),
             estimate_standardized = estimate / y_sd,
             se_standardized       = se / y_sd,
             CI_lower_standardized = estimate_standardized - qnorm(0.975) * se_standardized,
             CI_upper_standardized = estimate_standardized + qnorm(0.975) * se_standardized)

  } else { # exponentiate the effects
    results_dat <- results_dat %>% 
      mutate(exp_estimate = exp(estimate),
             exp_CIlower  = exp(CI_lower),
             exp_CIupper  = exp(CI_upper),
             y_sd = NA)
  }
  
  results_dat <- results_dat %>% 
    select(model, model_type, param_group, parameter, y_sd, estimate, CI_lower, CI_upper, pvalue, contains("standardized"), contains("exp"))
  row.names(results_dat) <- NULL
  
  return(results_dat)
})
results_dat <- modelResults_list %>% bind_rows() %>% 
  mutate(model       = factor(model, levels = names(model_names)),
         model_group = case_when(model %in% names(SCFA_markers) ~ "scfa",
                                 TRUE                           ~ "bcfa"),
         model_group = factor(model_group, levels = names(col_vector)))
```


## Intervention effects

```{r}
#| fig.width: 8
#| fig.height: 5

# 1) effects of Gaussian models
ylim_gaussianInt <- c(-.5, .5)
plot_dat_gaussian <- results_dat %>% 
  filter(model_type  == "Gaussian",
         param_group == "intervention") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")))
gg_gaussian <- plot_dat_gaussian %>% 
  ggplot(aes(x = model, y = estimate_standardized, color = model_group)) +
  geom_hline(yintercept = 0, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = CI_lower_standardized, ymax = CI_upper_standardized)) +
  geom_point() +
  scale_y_continuous("standardized estimate", limits = ylim_gaussianInt,
                     breaks = c(-.4,-.2,0,.2,.4),
                     labels = c("-0.4", "-0.2", "0", "+0.2", "+0.4")) +
  scale_color_manual(values = col_vector) +
  facet_wrap(~ parameter) +
  ggtitle("Intervention effects in Gaussian models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "none")
gg_gaussian


# 2) effects of log Gaussian and Gamma models
ylim_logInt <- c(.6,1.4)
plot_dat_exp <- results_dat %>% 
  filter(grepl("acid", model),
         model_type  != "Gaussian",
         param_group == "intervention") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")))
gg_exp <- plot_dat_exp %>% 
  ggplot(aes(x = model, y = exp_estimate, color = model_group)) +
  geom_hline(yintercept = 1, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = exp_CIlower, ymax = exp_CIupper)) +
  geom_point() +
  scale_y_continuous("estimate, on log2 scale", trans = "log2", limits = ylim_logInt,
                     breaks = c(.6, .8, 1, 1.2, 1.4),
                     labels = c("-40%", "-20%", "0", "+20%", "+40%"),
                     minor_breaks = c(.7, .9, 1.1, 1.3)) +
  scale_color_manual(values = col_vector) +
  facet_wrap(~ parameter) +
  ggtitle("Intervention effects in log Gaussian and Gamma models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "none")
gg_exp
```


## Phase, gender, age and BMI effects

```{r}
#| fig.width: 8
#| fig.height: 5

# 1) effects of Gaussian models
ylim_gaussian <- c(-1,1) * max(abs(c(results_dat$CI_lower_standardized[results_dat$parameter != "(Intercept)"],
                                     results_dat$CI_upper_standardized[results_dat$parameter != "(Intercept)"])), na.rm = TRUE)
results_dat %>% 
  filter(model_type  == "Gaussian",
         param_group %in% c("phase", "gender", "age", "BMI")) %>% 
  mutate(parameter = factor(parameter, levels = c("phasephase 2", "genderW", "age_centered50", "bmi_centered25"))) %>% 
  ggplot(aes(x = model, y = estimate_standardized, color = model_group)) +
  geom_hline(yintercept = 0, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = CI_lower_standardized, ymax = CI_upper_standardized)) +
  geom_point() +
  ylim(ylim_gaussian) +
  scale_color_manual(values = col_vector) +
  facet_wrap(~ parameter) +
  ggtitle("Covariate effects in Gaussian models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x     = element_text(hjust = 1, angle = 45),
        axis.title.x    = element_blank(),
        legend.position = "none")


# 2) effects of log Gaussian and Gamma models
ylim_log <- c(min(results_dat$exp_CIlower[results_dat$parameter != "(Intercept)"], na.rm = TRUE),
              max(results_dat$exp_CIupper[results_dat$parameter != "(Intercept)"], na.rm = TRUE))
results_dat %>% 
  filter(model_type  != "Gaussian",
         param_group %in% c("phase", "gender", "age", "BMI")) %>% 
  mutate(parameter = factor(parameter, levels = c("phasephase 2", "genderW", "age_centered50", "bmi_centered25"))) %>% 
  ggplot(aes(x = model, y = exp_estimate, color = model_group)) +
  geom_hline(yintercept = 1, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = exp_CIlower, ymax = exp_CIupper)) +
  geom_point() +
  scale_y_continuous("exponentiated estimate on log2 scale", trans = "log2", limits = ylim_log) +
  scale_color_manual(values = col_vector) +
  facet_wrap(~ parameter) +
  ggtitle("Covariate effects in log Gaussian and Gamma models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x     = element_text(hjust = 1, angle = 45),
        axis.title.x    = element_blank(),
        legend.position = "none")
```

## Interactions: Intervention by sex

```{r}
# create a results table with all effects
modelResults_sexInt_list <- lapply(names(model_names), function(x) {

  variable         <- model_names[x] %>% unname()
  model            <- model_sexInt_list[[x]][[1]]
  model_changedRef <- model_sexInt_list[[x]][[2]]
  model_type       <- modelType_list[[x]]

  log_wasApplied <- (model_type %in% c("Gamma", "log Gaussian"))

  results_dat1 <- summary(model)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(gender = levels(dat_model$gender)[1])
  results_dat2 <- summary(model_changedRef)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(gender = levels(dat_model$gender)[2])
  results_dat <- results_dat1 %>% dplyr::bind_rows(results_dat2) %>% 
    dplyr::rename(estimate   = "Estimate",
                  se         = "Std. Error",
                  pvalue     = "Pr(>|t|)") %>%
    mutate(model      = x,
           model_type = model_type,
           CI_lower   = estimate - qnorm(0.975) * se,
           CI_upper   = estimate + qnorm(0.975) * se)

  if (!log_wasApplied) { # standardize the effects by the standard deviation of the respnse variable
    results_dat <- results_dat %>%
      mutate(y_sd = sd(dat_model[[variable]], na.rm = TRUE),
             estimate_standardized = estimate / y_sd,
             se_standardized       = se / y_sd,
             CI_lower_standardized = estimate_standardized - qnorm(0.975) * se_standardized,
             CI_upper_standardized = estimate_standardized + qnorm(0.975) * se_standardized)

  } else { # exponentiate the effects
    results_dat <- results_dat %>%
      mutate(exp_estimate = exp(estimate),
             exp_CIlower  = exp(CI_lower),
             exp_CIupper  = exp(CI_upper))
  }

  results_dat <- results_dat %>%
    select(model, model_type, gender, parameter, estimate, CI_lower, CI_upper, contains("standardized"), contains("exp"), pvalue)
  row.names(results_dat) <- NULL

  return(results_dat)
})
results_dat_sexInt <- modelResults_sexInt_list %>% bind_rows() %>%
  mutate(model       = factor(model, levels = names(model_names)),
         model_group = case_when(model %in% names(SCFA_markers) ~ "scfa",
                                 TRUE                           ~ "bcfa"),
         model_group  = factor(model_group, levels = names(col_vector)),
         xCoord_shift = case_when(gender == min(gender) ~ -0.15,
                                  TRUE                  ~ 0.15))
```

```{r}
#| fig.width: 8
#| fig.height: 5

# 1) effects of Gaussian models
plot_dat <- results_dat_sexInt %>% 
  filter(model_type  == "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_gaussianInt <- c(-.8, .8)
gg_gaussian_sexInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = estimate_standardized, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 0, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = CI_lower_standardized, ymax = CI_upper_standardized, shape = gender), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("standardized estimate", limits = ylim_gaussianInt,
                     breaks = c(-.8,-.4,0,.4,.8),
                     labels = c("-0.8", "-0.4", "0", "+0.4", "+0.8")) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("sex", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("sex-specific Intervention effects in Gaussian models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_gaussian_sexInt


# 2) effects of log Gaussian and Gamma models
plot_dat <- results_dat_sexInt %>% 
  filter(model_type  != "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_logInt <- c(.5,1.6)
gg_exp_sexInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = exp_estimate, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 1, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = exp_CIlower, ymax = exp_CIupper, shape = gender), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("estimate, on log2 scale", trans = "log2", limits = ylim_logInt,
                     breaks = c(.6, .8, 1, 1.2, 1.4, 1.6),
                     labels = c("-40%", "-20%", "0", "+20%", "+40%", "+60%"),
                     minor_breaks = c(.5, .7, .9, 1.1, 1.3, 1.5)) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("sex", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("sex-specific Intervention effects in log Gaussian and Gamma models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_exp_sexInt
```


## Interactions: Intervention by age

```{r}
# create a results table with all effects
modelResults_ageInt_list <- lapply(names(model_names), function(x) {

  variable         <- model_names[x] %>% unname()
  model            <- model_ageInt_list[[x]][[1]]
  model_changedRef <- model_ageInt_list[[x]][[2]]
  model_type       <- modelType_list[[x]]

  log_wasApplied <- (model_type %in% c("Gamma", "log Gaussian"))

  results_dat1 <- summary(model)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(age_group = levels(dat_model$age_cat)[1])
  results_dat2 <- summary(model_changedRef)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(age_group = levels(dat_model$age_cat)[2])
  results_dat <- results_dat1 %>% dplyr::bind_rows(results_dat2) %>% 
    dplyr::rename(estimate   = "Estimate",
                  se         = "Std. Error",
                  pvalue     = "Pr(>|t|)") %>%
    mutate(model      = x,
           model_type = model_type,
           CI_lower   = estimate - qnorm(0.975) * se,
           CI_upper   = estimate + qnorm(0.975) * se)

  if (!log_wasApplied) { # standardize the effects by the standard deviation of the respnse variable
    results_dat <- results_dat %>%
      mutate(y_sd = sd(dat_model[[variable]], na.rm = TRUE),
             estimate_standardized = estimate / y_sd,
             se_standardized       = se / y_sd,
             CI_lower_standardized = estimate_standardized - qnorm(0.975) * se_standardized,
             CI_upper_standardized = estimate_standardized + qnorm(0.975) * se_standardized)

  } else { # exponentiate the effects
    results_dat <- results_dat %>%
      mutate(exp_estimate = exp(estimate),
             exp_CIlower  = exp(CI_lower),
             exp_CIupper  = exp(CI_upper))
  }

  results_dat <- results_dat %>%
    select(model, model_type, age_group, parameter, estimate, CI_lower, CI_upper, contains("standardized"), contains("exp"), pvalue)
  row.names(results_dat) <- NULL

  return(results_dat)
})
results_dat_ageInt <- modelResults_ageInt_list %>% bind_rows() %>%
  mutate(model       = factor(model, levels = names(model_names)),
         model_group = case_when(model %in% names(SCFA_markers) ~ "scfa",
                                 TRUE                           ~ "bcfa"),
         model_group  = factor(model_group, levels = names(col_vector)),
         xCoord_shift = case_when(age_group == min(age_group) ~ -0.15,
                                  TRUE                        ~ 0.15))
```

```{r}
#| fig.width: 8
#| fig.height: 5

# 1) effects of Gaussian models
plot_dat <- results_dat_ageInt %>% 
  filter(model_type  == "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_gaussianInt <- c(-.8, .8)
gg_gaussian_ageInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = estimate_standardized, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 0, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = CI_lower_standardized, ymax = CI_upper_standardized, shape = age_group), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("standardized estimate", limits = ylim_gaussianInt,
                     breaks = c(-.8,-.4,0,.4,.8),
                     labels = c("-0.8", "-0.4", "0", "+0.4", "+0.8")) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("age group", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("age-specific Intervention effects in Gaussian models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_gaussian_ageInt



# 2) effects of log Gaussian and Gamma models
plot_dat <- results_dat_ageInt %>% 
  filter(model_type  != "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_logInt <- c(.5,1.5)
gg_exp_ageInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = exp_estimate, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 1, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = exp_CIlower, ymax = exp_CIupper, shape = age_group), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("estimate, on log2 scale", trans = "log2", limits = ylim_logInt,
                     breaks = c(.6, .8, 1, 1.2, 1.4),
                     labels = c("-40%", "-20%", "0", "+20%", "+40%"),
                     minor_breaks = c(.5, .7, .9, 1.1, 1.3, 1.5)) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("age group", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("age-specific Intervention effects in log Gaussian and Gamma models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_exp_ageInt
```


## Interactions: Intervention by BMI

```{r}
# create a results table with all effects
modelResults_bmiInt_list <- lapply(names(model_names), function(x) {
  
  variable         <- model_names[x] %>% unname()
  model            <- model_bmiInt_list[[x]][[1]]
  model_changedRef <- model_bmiInt_list[[x]][[2]]
  model_type       <- modelType_list[[x]]

  log_wasApplied <- (model_type %in% c("Gamma", "log Gaussian"))

  results_dat1 <- summary(model)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(bmi_group = levels(dat_model$bmi_cat)[1])
  results_dat2 <- summary(model_changedRef)$p.table %>% as.data.frame() %>%
    mutate(parameter   = row.names(.)) %>%
    filter(grepl("intervention", parameter),
           !grepl(":", parameter)) %>% 
    mutate(bmi_group = levels(dat_model$bmi_cat)[2])
  results_dat <- results_dat1 %>% dplyr::bind_rows(results_dat2) %>% 
    dplyr::rename(estimate   = "Estimate",
                  se         = "Std. Error",
                  pvalue     = "Pr(>|t|)") %>%
    mutate(model      = x,
           model_type = model_type,
           CI_lower   = estimate - qnorm(0.975) * se,
           CI_upper   = estimate + qnorm(0.975) * se)

  if (!log_wasApplied) { # standardize the effects by the standard deviation of the respnse variable
    results_dat <- results_dat %>%
      mutate(y_sd = sd(dat_model[[variable]], na.rm = TRUE),
             estimate_standardized = estimate / y_sd,
             se_standardized       = se / y_sd,
             CI_lower_standardized = estimate_standardized - qnorm(0.975) * se_standardized,
             CI_upper_standardized = estimate_standardized + qnorm(0.975) * se_standardized)

  } else { # exponentiate the effects
    results_dat <- results_dat %>%
      mutate(exp_estimate = exp(estimate),
             exp_CIlower  = exp(CI_lower),
             exp_CIupper  = exp(CI_upper))
  }

  results_dat <- results_dat %>%
    select(model, model_type, bmi_group, parameter, estimate, CI_lower, CI_upper, contains("standardized"), contains("exp"), pvalue)
  row.names(results_dat) <- NULL

  return(results_dat)
})
results_dat_bmiInt <- modelResults_bmiInt_list %>% bind_rows() %>%
  mutate(model       = factor(model, levels = names(model_names)),
         model_group = case_when(model %in% names(SCFA_markers) ~ "scfa",
                                 TRUE                           ~ "bcfa"),
         model_group  = factor(model_group, levels = names(col_vector)),
         xCoord_shift = case_when(bmi_group == min(bmi_group) ~ -0.15,
                                  TRUE                        ~ 0.15))
```

```{r}
#| fig.width: 8
#| fig.height: 5

# 1) effects of Gaussian models
plot_dat <- results_dat_bmiInt %>% 
  filter(model_type  == "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_gaussianInt <- c(-.8, .8)
gg_gaussian_bmiInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = estimate_standardized, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 0, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = CI_lower_standardized, ymax = CI_upper_standardized, shape = bmi_group), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("standardized estimate", limits = ylim_gaussianInt,
                     breaks = c(-.8,-.4,0,.4,.8),
                     labels = c("-0.8", "-0.4", "0", "+0.4", "+0.8")) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("BMI group", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("BMI-specific Intervention effects in Gaussian models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_gaussian_bmiInt



# 2) effects of log Gaussian and Gamma models
plot_dat <- results_dat_bmiInt %>% 
  filter(model_type  != "Gaussian") %>% 
  mutate(parameter = factor(parameter, levels = c("interventionFresh", "interventionPasteurized"),
                            labels = c("fresh intervention", "pasteurized intervention")),
         model     = droplevels(model),
         model     = factor(model, levels = levels(model),
                            labels = levels(model) %>% gsub(pattern = " \\(Stool)", replacement = "")),
         model_xCoord = as.numeric(model) + xCoord_shift)

ylim_logInt <- c(.5,2)
gg_exp_bmiInt <- plot_dat %>% 
  filter(grepl("acid", model)) %>% 
  ggplot(aes(x = model_xCoord, y = exp_estimate, color = model_group, fill = model_group)) +
  geom_hline(yintercept = 1, linetype = 2, col = "gray30") +
  geom_pointrange(aes(ymin = exp_CIlower, ymax = exp_CIupper, shape = bmi_group), size = 0.25) +
  scale_x_continuous(breaks = 1:length(levels(plot_dat$model)), labels = levels(plot_dat$model)) +
  scale_y_continuous("estimate, on log2 scale", trans = "log2", limits = ylim_logInt,
                     breaks = c(.6, .8, 1, 1.2, 1.4, 1.6, 1.8, 2),
                     labels = c("-40%", "-20%", "0", "+20%", "+40%", "+60%", "+80%", "+100%"),
                     minor_breaks = c(.5, .7, .9, 1.1, 1.3, 1.5, 1.7, 1.9)) +
  scale_color_manual(values = col_vector, guide = "none") +
  scale_fill_manual(values = col_vector, guide = "none") +
  scale_shape_manual("BMI group", values = c(25,24)) +
  facet_wrap(~ parameter) +
  ggtitle("BMI-specific Intervention effects in log Gaussian and Gamma models", "with uncorrected 95% confidence intervals") +
  theme(axis.text.x        = element_text(hjust = 1, angle = 45),
        axis.title.x       = element_blank(),
        panel.grid.minor.y = element_line(color = "gray96"),
        panel.grid.major.x = element_blank(),
        legend.position    = "bottom")
gg_exp_bmiInt
```


# Corrected p-values for intervention effects

Correct for multiple testing by applying the Benjamini-Hochberg correction to all intervention effects.

```{r}
results_dat_interventions <- results_dat %>% 
  filter(grepl("intervention", parameter)) %>% 
  mutate(pvalue_corrected   = p.adjust(pvalue, method = "BH") %>% round(4),
         pvalue_significant = ifelse(pvalue_corrected < 0.1, "yes", "no"))

results_dat_interventions %>% 
  select(model, model_type, parameter, pvalue_corrected, pvalue_significant) %>% 
  kable() %>% 
  kable_styling()
```


# Corrected p-values for interaction effects

For the interaction effects, correct for multiple testing by applying the Benjamini-Hochberg correction.
The correction is applied individually, once for the sex / the age / the BMI interaction models.

```{r}
# sex
results_dat_sexInt <- results_dat_sexInt %>% 
  mutate(pvalue_corrected   = p.adjust(pvalue, method = "BH") %>% round(4),
         pvalue_significant = ifelse(pvalue_corrected < 0.1, "yes", "no"))

results_dat_sexInt %>% 
  select(model, model_type, gender, parameter, pvalue_corrected, pvalue_significant) %>% 
  kable() %>% 
  kable_styling()

# age
results_dat_ageInt <- results_dat_ageInt %>% 
  mutate(pvalue_corrected   = p.adjust(pvalue, method = "BH") %>% round(4),
         pvalue_significant = ifelse(pvalue_corrected < 0.1, "yes", "no"))

results_dat_ageInt %>% 
  select(model, model_type, age_group, parameter, pvalue_corrected, pvalue_significant) %>% 
  kable() %>% 
  kable_styling()

# BMI
results_dat_bmiInt <- results_dat_bmiInt %>% 
  mutate(pvalue_corrected   = p.adjust(pvalue, method = "BH") %>% round(4),
         pvalue_significant = ifelse(pvalue_corrected < 0.1, "yes", "no"))

results_dat_bmiInt %>% 
  select(model, model_type, bmi_group, parameter, pvalue_corrected, pvalue_significant) %>% 
  kable() %>% 
  kable_styling()
```


# Figures for publication

```{r}
#| column: page
#| fig-width: 16

# save the results datasets, to be able to jointly plot blood and stool results in the blood document
# write.csv(results_dat, file = "9_resultsDat_stoolSCFAmodels.csv", row.names = FALSE)
# write.csv(results_dat_ageInt, file = "9_resultsDat_ageInt_stoolSCFAmodels.csv", row.names = FALSE)
# write.csv(results_dat_sexInt, file = "9_resultsDat_sexInt_stoolSCFAmodels.csv", row.names = FALSE)
# write.csv(results_dat_bmiInt, file = "9_resultsDat_bmiInt_stoolSCFAmodels.csv", row.names = FALSE)

# 1) main model results
(gg_gaussian + gg_exp) + patchwork::plot_layout(nrow = 1, widths = c(.33,.67))
# ggsave("modelResults_stoolSCFAs.png", width = 13, height = 4)

# 2) interactions with gender
(gg_gaussian_sexInt + gg_exp_sexInt) + patchwork::plot_layout(nrow = 1, widths = c(.33,.67))
# ggsave("modelResults_sexInteraction_stoolSCFAs.png", width = 14, height = 4)

# 3) interactions with age
(gg_gaussian_ageInt + gg_exp_ageInt) + patchwork::plot_layout(nrow = 1, widths = c(.33,.67))
# ggsave("modelResults_ageInteraction_stoolSCFAs.png", width = 14, height = 4)

# 4) interactions with age
(gg_gaussian_bmiInt + gg_exp_bmiInt) + patchwork::plot_layout(nrow = 1, widths = c(.33,.67))
# ggsave("modelResults_bmiInteraction_stoolSCFAs.png", width = 14, height = 4)
```


# Descriptive response patterns: Compare how the target parameters change per person

## SCFA / BCFA

```{r}
#| column: page

# per person and period, add a variable indicating how the markers changed
dat_change <- dat_model %>% 
  arrange(participant_id, timepoint) %>% 
  group_by(participant_id, period) %>% 
  filter(n() == 2) %>% 
  mutate(methylbutyric_acid2_diff = diff(stool_methylbutyricAcid),
         acetic_acid_diff         = diff(stool_aceticAcid),
         butyric_acid_diff        = diff(stool_butyricAcid),
         hexanoic_acid_diff       = diff(stool_hexanoicAcid),
         isobutyric_acid_diff     = diff(stool_isobutyricAcid),
         isovaleric_acid_diff     = diff(stool_isovalericAcid),
         propionic_acid_diff      = diff(stool_propionicAcid),
         valeric_acid_diff        = diff(stool_valericAcid)) %>% 
  slice(2) %>% 
  ungroup() %>% 
  mutate(MBacid    = case_when(methylbutyric_acid2_diff >= 0 ~ "bad (increasing)",
                               methylbutyric_acid2_diff <  0 ~ "good (decreasing)"),
         ACacid    = case_when(acetic_acid_diff         >= 0 ~ "bad (increasing)",
                               acetic_acid_diff         <  0 ~ "good (decreasing)"),
         BTacid    = case_when(butyric_acid_diff        >= 0 ~ "bad (increasing)",
                               butyric_acid_diff        <  0 ~ "good (decreasing)"),
         HXacid    = case_when(hexanoic_acid_diff       >= 0 ~ "bad (increasing)",
                               hexanoic_acid_diff       <  0 ~ "good (decreasing)"),
         isoBTacid = case_when(isobutyric_acid_diff     >= 0 ~ "bad (increasing)",
                               isobutyric_acid_diff     <  0 ~ "good (decreasing)"),
         isoVLacid = case_when(isovaleric_acid_diff     >= 0 ~ "bad (increasing)",
                               isovaleric_acid_diff     <  0 ~ "good (decreasing)"),
         PPacid    = case_when(propionic_acid_diff      >= 0 ~ "bad (increasing)",
                               propionic_acid_diff      <  0 ~ "good (decreasing)"),
         VLacid    = case_when(valeric_acid_diff        >= 0 ~ "bad (increasing)",
                               valeric_acid_diff        <  0 ~ "good (decrasing)"))


# add a variable defining the change pattern, and create a frequency table of it
tab_change <- dat_change %>% 
  mutate(acid_changePattern = paste(MBacid, ACacid, BTacid, HXacid, isoBTacid, isoVLacid, PPacid, VLacid)) %>% 
  group_by(acid_changePattern) %>% 
  summarize(MBacid              = first(MBacid),
            ACacid              = first(ACacid),
            BTacid              = first(BTacid),
            HXacid              = first(HXacid),
            isoBTacid           = first(isoBTacid),
            isoVLacid           = first(isoVLacid),
            PPacid              = first(PPacid),
            VLacid              = first(VLacid),
            N_freshIntervention = sum(intervention == "Fresh"),
            N_pastIntervention  = sum(intervention == "Pasteurized")) %>% 
  ungroup() %>% 
  mutate(fake_rowGrouping = 1:nrow(.)) %>% 
  group_by(fake_rowGrouping) %>% 
  mutate(pattern = case_when(length(strsplit(acid_changePattern, "good")[[1]]) == 9 ~ "8/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 8 ~ "7/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 7 ~ "6/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 6 ~ "5/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 5 ~ "4/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 4 ~ "3/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 3 ~ "2/8 good",
                             length(strsplit(acid_changePattern, "good")[[1]]) == 2 ~ "1/8 good",
                             TRUE                                                   ~ "0/8 good"),
         pattern = factor(pattern, levels = c("8/8 good", "7/8 good", "6/8 good", "5/8 good", "4/8 good",
                                              "3/8 good", "2/8 good", "1/8 good", "0/8 good"))) %>% 
  ungroup() %>% 
  arrange(pattern, acid_changePattern) %>% 
  select(-acid_changePattern, -fake_rowGrouping) %>% 
  select(pattern, everything())
  
# print the table (summarized form)
tab_change %>% 
  group_by(pattern) %>% 
  summarize(N_freshIntervention = sum(N_freshIntervention),
            N_pastIntervention  = sum(N_pastIntervention)) %>% 
  kable() %>% 
  kable_styling()

# print the table (detailed form)
tab_change %>% 
  mutate(MBacid    = cell_spec(MBacid,    background = ifelse(grepl("good", MBacid),    "#92C7CF", "#E6A4B4")),
         ACacid    = cell_spec(ACacid,    background = ifelse(grepl("good", ACacid),    "#92C7CF", "#E6A4B4")),
         BTacid    = cell_spec(BTacid,    background = ifelse(grepl("good", BTacid),    "#92C7CF", "#E6A4B4")),
         HXacid    = cell_spec(HXacid,    background = ifelse(grepl("good", HXacid),    "#92C7CF", "#E6A4B4")),
         isoBTacid = cell_spec(isoBTacid, background = ifelse(grepl("good", isoBTacid), "#92C7CF", "#E6A4B4")),
         isoVLacid = cell_spec(isoVLacid, background = ifelse(grepl("good", isoVLacid), "#92C7CF", "#E6A4B4")),
         PPacid    = cell_spec(PPacid,    background = ifelse(grepl("good", PPacid),    "#92C7CF", "#E6A4B4")),
         VLacid    = cell_spec(VLacid,    background = ifelse(grepl("good", VLacid),    "#92C7CF", "#E6A4B4"))) %>% 
  kable(escape = FALSE) %>% 
  kable_styling()
```
