---
title: "How to use sauerkrautTaxonomyBuddy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HowToUseSauerkrautBuddy}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| warning: false

library(sauerkrautTaxonomyBuddy)
library(mia)                  # microbiome analysis
library(SummarizedExperiment) # microbiome analysis
library(scater)               # dimension reduction

library(dplyr) # data handling

# set ggplot theme
theme_set(theme_minimal())
```

```{r set file paths}
define_dataPaths()
```

## Main functions for taxonomic analysis

### Create a tse object from all taxonomic analysis data files

```{r}
tse <- readAndPrepare_mainTaxonomicData(path_groupInfoRdata               = path_groupInfoRdata,
                                        path_participantRdata             = path_participantRdata,
                                        path_samplesIDxlsx                = path_samplesIDxlsx,
                                        path_dietInfoCsv                  = path_dietInfoCsv,
                                        path_nutrientInfoCsv              = path_nutrientInfoCsv,
                                        path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
                                        path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
                                        path_T4questXlsx                  = path_T4questXlsx,
                                        path_stoolInfoXlsx                = path_stoolInfoXlsx,
                                        path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
                                        path_absTaxonomyData              = path_absTaxonomyData,
                                        path_relTaxonomyData              = path_relTaxonomyData,
                                        path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                        path_krautLookupXlsx              = path_krautLookupXlsx,
                                        path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                        path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                        path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                        path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                        aggregation_level                 = "species",
                                        exclude_sickParticipants          = "taxonomy paper")

# add further features to colData(tse)
tse <- tse_addTaxaFeatures(tse          = tse,
                           assay_relAbd = "relAbd")


# how to access the individual components of the data object:
# 1. absolute abundances
assay(tse, "absAbd")
# 2. relative abundances
assay(tse, "relAbd")
# 3. basic experiment and subject data on each sample, and calculated statistics
colData(tse)
# 4. taxonomic tree information for each bacteria species / strain
rowData(tse)
```


### Calculate Bray-Curtis and UniFrac distances

```{r}
datDist_list <- calculate_distances(tse, "relAbd")
```


### Crossover design boxplots

```{r}
dat_crossoverPlots <- tse %>% colData() %>% as.data.frame() %>% 
  filter(timepoint != "T5")

plot_crossoverBoxplots(dat              = dat_crossoverPlots,
                       y_var            = "richness_observed",
                       intervention_var = "intervention",
                       t_var            = "treatment",
                       subject_var      = "participant_id",
                       title            = "Observed richness index")
```


### Dimension reduction plots

```{r}
#| warning: false

# estimate UMAP as a dimension reduction technique
tse <- tse %>% scater::runUMAP(name = "UMAP", exprs_values = "relAbd")

# create plot list
gg_list <- plot_dimReduction(tse               = tse,
                             dimReduction_name = "UMAP",
                             assay             = "relAbd",
                             taxRank_dominance = "family")

# one exemplaric plot
gg_list$gg_interventionTimes
```



## Main functions for functional analysis data

### Create a tse object from all functional analysis pathway data files

```{r}
tse_PW <- readAndPrepare_mainFunctionalPathwayData(
  path_KEGG_KO                      = path_KEGGKO_funData,
  path_KEGG_pathways                = path_KEGGPathways_funData,
  path_groupInfoRdata               = path_groupInfoRdata,
  path_participantRdata             = path_participantRdata,
  path_samplesIDxlsx                = path_samplesIDxlsx,
  path_dietInfoCsv                  = path_dietInfoCsv,
  path_nutrientInfoCsv              = path_nutrientInfoCsv,
  path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
  path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
  path_T4questXlsx                  = path_T4questXlsx,
  path_stoolInfoXlsx                = path_stoolInfoXlsx,
  path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
  path_sampleLookupXlsx             = path_sampleLookupXlsx,
  path_krautLookupXlsx              = path_krautLookupXlsx,
  path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
  path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
  path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
  path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
  exclude_sickParticipants          = "taxonomy paper"
)


# how to access the individual components of the data object:
# 1. absolute abundances
assay(tse_PW, "tpmAbd")
# 2. relative abundances
assay(tse_PW, "relAbd")
# 3. basic experiment and subject data on each sample, and calculated statistics
colData(tse_PW)
# 4. pathway information for each function
rowData(tse_PW)
```


### Create a tse object from all functional analysis KO data files

```{r}
tse_KO <- readAndPrepare_mainFunctionalKOdata(
  path_KEGG_KO                      = path_KEGGKO_funData,
  path_KEGG_pathways                = path_KEGGPathways_funData,
  path_groupInfoRdata               = path_groupInfoRdata,
  path_participantRdata             = path_participantRdata,
  path_samplesIDxlsx                = path_samplesIDxlsx,
  path_dietInfoCsv                  = path_dietInfoCsv,
  path_nutrientInfoCsv              = path_nutrientInfoCsv,
  path_bloodMetabolomeXlsx          = path_bloodMetabolomeXlsx,
  path_stoolMetabolomeXlsx          = path_stoolMetabolomeXlsx,
  path_T4questXlsx                  = path_T4questXlsx,
  path_stoolInfoXlsx                = path_stoolInfoXlsx,
  path_bodyMeasuresXlsx             = path_bodyMeasuresXlsx,
  path_sampleLookupXlsx             = path_sampleLookupXlsx,
  path_krautLookupXlsx              = path_krautLookupXlsx,
  path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
  path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
  path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
  path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
  exclude_sickParticipants          = "taxonomy paper"
)


# how to access the individual components of the data object:
# 1. absolute abundances
assay(tse_KO, "tpmAbd")
# 2. relative abundances
assay(tse_KO, "relAbd")
# 3. basic experiment and subject data on each sample, and calculated statistics
colData(tse_KO)
# 4. KO information
rowData(tse_KO)
```



## Functions for reading individual datasets

### Basic experiment data

```{r}
dat_exp <- readAndPrepare_experimentData(path_groupInfoRdata      = path_groupInfoRdata,
                                         path_participantRdata    = path_participantRdata,
                                         path_samplesIDxlsx       = path_samplesIDxlsx,
                                         path_dietInfoCsv         = path_dietInfoCsv,
                                         path_nutrientInfoCsv     = path_nutrientInfoCsv,
                                         path_bloodMetabolomeXlsx = path_bloodMetabolomeXlsx,
                                         path_stoolMetabolomeXlsx = path_stoolMetabolomeXlsx,
                                         path_T4questXlsx         = path_T4questXlsx,
                                         path_stoolInfoXlsx       = path_stoolInfoXlsx,
                                         path_bodyMeasuresXlsx    = path_bodyMeasuresXlsx,
                                         path_sampleLookupXlsx    = path_sampleLookupXlsx,
                                         path_krautLookupXlsx     = path_krautLookupXlsx,
                                         exclude_sickParticipants = "taxonomy paper")

dat_lookupList <- readAndPrepare_lookupData(path_sampleLookupXlsx    = path_sampleLookupXlsx,
                                            path_krautLookupXlsx     = path_krautLookupXlsx,
                                            exclude_sickParticipants = "taxonomy paper")
dat_lookupStoolTax <- datList_tax$dat_lookupStool
dat_lookupKrautTax <- datList_tax$dat_lookupKraut
rm(dat_lookupList)
```


### Symptom diary data

```{r}
dat_symp <- readAndPrepare_symptomDiaries(path_symptomDiaryData)
```


### Taxonomy data

```{r}
datList_tax <- readAndPrepare_taxonomy(path_taxonomyData         = path_relTaxonomyData,
                                       path_sampleLookupXlsx     = path_sampleLookupXlsx,
                                       path_krautLookupXlsx      = path_krautLookupXlsx,
                                       aggregation_level         = "species",
                                       exclude_sickParticipants  = "taxonomy paper")
dat_stoolTax       <- datList_tax$dat_stoolSamples
dat_stoolDuplTax   <- datList_tax$dat_stoolDuplicateSamples
dat_krautTax       <- datList_tax$dat_krautSamples
dat_lookupStoolTax <- datList_tax$dat_lookupStool
dat_lookupKrautTax <- datList_tax$dat_lookupKraut
rm(datList_tax)
```


### Functional analysis data

```{r}
# prepare relative abundances
datList_fun <- readAndPrepare_functionalData(path_KEGG_KO             = path_KEGGKO_funData,
                                             path_KEGG_pathways       = path_KEGGPathways_funData,
                                             path_sampleLookupXlsx    = path_sampleLookupXlsx,
                                             path_krautLookupXlsx     = path_krautLookupXlsx,
                                             abundance_unit           = "TPM",
                                             exclude_sickParticipants = "taxonomy paper")
dat_funKO_stoolSamples <- datList_fun$datKO_stoolSamples
dat_funPW_stoolSamples <- datList_fun$datPathways_stoolSamples
```


### Blood marker data

```{r}
datList_blood <- readAndPrepare_bloodMarkerData(path_bloodMarkerBostonRdata       = path_bloodMarkerBostonRdata,
                                                path_bloodMarkerZentrallaborRdata = path_bloodMarkerZentrallaborRdata,
                                                path_bloodMarkerBostonXlsx        = path_bloodMarkerBostonXlsx,
                                                path_bloodMarkerZentrallaborXlsx  = path_bloodMarkerZentrallaborXlsx,
                                                path_sampleLookupXlsx             = path_sampleLookupXlsx,
                                                path_krautLookupXlsx              = path_krautLookupXlsx,
                                                exclude_sickParticipants          = "taxonomy paper")
dat_blood            <- datList_blood$main_data
dat_blood_BostonDupl <- datList_blood$duplicates_BostonData
dat_blood_zlDupl     <- datList_blood$duplicates_ZentrallaborData
```
